---
title: "Complete Workflow with hbsaems 1.0.0"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete Workflow with hbsaems 1.0.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

```{r load-libs, eval = TRUE, message = FALSE}
library(hbsaems)
```

A typical Bayesian SAE workflow with `hbsaems` v1.0.0 follows seven steps,
each backed by a single primary function.

## See also

For deeper coverage of topics beyond this introductory workflow, see the
**articles on the package website**:

<https://madsyair.github.io/hbsaems/articles/>

The articles cover:

- Distribution-specific walkthroughs (Beta logit-normal, binomial
  logit-normal, lognormal-lognormal)
- Spatial models (CAR / SAR / BYM2)
- Handling missing data (deletion, multiple imputation, joint Bayesian
  imputation)
- The interactive Shiny dashboard
- Advanced features (custom families, benchmarking, smooth terms)
- AST-based formula manipulation (internals)
- Migration guide for users coming from earlier versions


## 1. Prior predictive check

```r
data("data_fhnorm")
model_prior <- hbm(
  formula           = brms::bf(y ~ x1 + x2 + x3),
  data              = data_fhnorm,
  re                = ~ (1 | regency),    # area-level random effect (Fay-Herriot)
  sampling_variance = "D",                 # KNOWN sampling variance D_i
  sample_prior      = "only",
  prior             = c(
    brms::prior(normal(0, 1),  class = "b"),
    brms::prior(normal(10, 5), class = "Intercept"),
    brms::prior(normal(0, 2),  class = "sd")
  ),
  chains = 2, iter = 1000
)
pc <- prior_check(model_prior,
                  data         = data_fhnorm,
                  response_var = "y")
plot(pc)
```

The `sampling_variance = "D"` argument is the **defining feature** of a
Fay-Herriot model: the sampling variance \eqn{D_i} is treated as
\strong{known} from the design (it is not estimated), so brms pins the
observation-level \eqn{\sigma_i} to \eqn{\sqrt{D_i}}.  Omitting this
argument makes the model unidentified because the residual \eqn{\sigma}
and the area-RE \eqn{\sigma_u} would compete to explain the same
variance, typically producing divergent transitions and very wide
credible intervals.

## 2. Fit the model

```r
model <- hbm(
  formula           = brms::bf(y ~ x1 + x2 + x3),
  data              = data_fhnorm,
  re                = ~ (1 | regency),       # area-level random effect (Fay-Herriot)
  sampling_variance = "D",                    # known sampling variance from survey
  control           = list(adapt_delta = 0.99),
  chains            = 4, iter = 4000, warmup = 2000, cores = 2
)
summary(model)
```

Two settings deserve attention:

* **`sampling_variance = "D"`** -- the column \code{D} in
  \code{data_fhnorm} holds the known sampling variance \eqn{D_i} for
  each area; \pkg{hbsaems} translates this into an offset on the
  observation-level standard deviation so brms / Stan never tries to
  estimate it.
* **`adapt_delta = 0.99`** -- Fay-Herriot likelihoods can produce
  mild funnel geometry between \eqn{\sigma_u} and the area random
  effects \eqn{u_i} when many areas have small \eqn{D_i}.  Raising
  \code{adapt_delta} from the brms default 0.95 to 0.99 buys a more
  conservative leapfrog step and eliminates the occasional divergent
  transition without re-parameterising the model.

## 3. Convergence diagnostics

```r
diag <- convergence_check(model)
is_converged(model)
summary(diag)
plot(diag, type = "trace")
hbm_warnings(model)
```

## 4. Goodness-of-fit

```r
gof <- model_compare(model)         # single-model: LOO, WAIC, pp_check
summary(gof)
plot(gof, type = "pp_check")
```

## 5. Compare alternatives

```r
model2 <- hbm(brms::bf(y ~ x1 + x2), data = data_fhnorm,
              re = ~ (1 | regency),
              sampling_variance = "D",
              control = list(adapt_delta = 0.99),
              chains = 4, iter = 4000)
model3 <- hbm(brms::bf(y ~ x1),      data = data_fhnorm,
              re = ~ (1 | regency),
              sampling_variance = "D",
              control = list(adapt_delta = 0.99),
              chains = 4, iter = 4000)

model_compare(model, model2)
model_compare_all(full = model, medium = model2, simple = model3)
```

## 6. Small-area estimates

```r
est <- sae_predict(model)
summary(est)
plot(est, type = "predictions")
plot(est, type = "uncertainty")

# Bayesian model averaging
avg <- model_average(model, model2, weights = c(0.7, 0.3))
```

## 7. Post-processing predictions

```r
log_est  <- sae_transform(est, log)
sc_est   <- sae_scale(est)
hi_est   <- sae_filter(est, est$pred > stats::median(est$pred))
agg_est  <- sae_aggregate(sae_predict(model),
                          sae_predict(model2),
                          method = "mean")
```
