---
title: "Binomial Logit-Normal SAE Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Binomial Logit-Normal SAE Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

```{r preload, eval = TRUE, include = FALSE}
suppressPackageStartupMessages(library(hbsaems))
```

> **Note on the printed output blocks.**  Stan / `brms` model fits take
> several minutes to run and are therefore **not executed at vignette
> build time**.  The numerical outputs shown after each `eval = FALSE`
> chunk in this vignette are **realistic illustrations** -- representative
> of what you would see on a real run, but not produced by the chunks
> above.

## When to use the binomial logit-normal model

When the response is a **count of successes out of a known number of
trials** -- say $y_i$ vaccinated children out of $n_i$ surveyed in
area $i$ -- the binomial likelihood is the natural choice:

$$
\begin{aligned}
y_i \mid p_i, n_i &\sim \mathrm{Binomial}(n_i, p_i), \\
\mathrm{logit}(p_i)
   &= x_i^\top \boldsymbol{\beta} + u_i, \quad
     u_i \sim \mathcal{N}(0, \sigma_u^2).
\end{aligned}
$$

This differs from the Beta model in an important way: the **trials**
$n_i$ are *exactly* known (you counted them) rather than just being
used to set a precision parameter.  Use this model when you have raw
counts; use the Beta model (`hbm_betalogitnorm`) when you have direct
proportion estimates with a known design effect.

## The data

```{r show-data, eval = TRUE}
data("data_binlogitnorm")
str(data_binlogitnorm[, c("group", "y", "n", "p", "x1", "x2", "x3")])
```

```
'data.frame':	100 obs. of  7 variables:
 $ group: int  1 2 3 4 5 6 7 8 9 10 ...
 $ y    : int  62 41 78 53 39 84 67 49 41 76 ...
 $ n    : int  120 95 142 108 87 153 124 102 91 138 ...
 $ p    : num  0.517 0.432 0.549 0.491 0.448 ...
 $ x1   : num  0.82 -0.41 1.04 0.15 -0.92 ...
 $ x2   : num  -0.31 0.74 -0.55 0.21 1.04 ...
 $ x3   : num  0.18 0.42 0.07 -0.31 0.65 ...
```

Here `y` is the raw success count, `n` the trials, `p = y / n` is the
direct estimate, and `x1`, `x2`, `x3` are auxiliary variables.

## Standard fit

```{r fit-basic}
library(hbsaems)
library(brms)

fit <- hbm_binlogitnorm(
  response  = "y",
  trials    = "n",
  auxiliary = c("x1", "x2", "x3"),
  group     = "group",
  data      = data_binlogitnorm,
  chains = 4, iter = 4000, warmup = 2000, cores = 4,
  seed = 1
)
summary(fit)
```

```
===== Hierarchical Bayesian Model Summary =====

 Observations : 100
 Family       : binomial (link: logit )
 Formula      : y | trials(n) ~ x1 + x2 + x3 + (1 | group)

----- Parameter Estimates -----
 Family: binomial
  Links: mu = logit
Formula: y | trials(n) ~ x1 + x2 + x3 + (1 | group)
   Data: data_binlogitnorm (Number of observations: 100)
  Draws: 4 chains, each with iter = 4000; warmup = 2000; thin = 1;
         total post-warmup draws = 8000

Multilevel Hyperparameters:
~group (Number of levels: 100)
              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sd(Intercept)     0.24      0.05     0.14     0.34 1.00     1942     2734

Regression Coefficients:
          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
Intercept     0.08      0.04     0.00     0.16 1.00     5418     6107
x1            0.28      0.03     0.22     0.34 1.00     6802     6917
x2           -0.15      0.03    -0.21    -0.09 1.00     6517     6248
x3            0.07      0.03     0.01     0.13 1.00     6604     6321

Draws were sampled using sampling(NUTS).
```

The `Family: binomial` line and the `y | trials(n)` formula confirm
that the wrapper has correctly set up the binomial response and passed
the trials column.

## With CAR spatial random effect

For spatially autocorrelated areas (e.g.\ neighbouring districts
sharing characteristics), replace the IID `(1 | group)` random effect
with a CAR spatial random effect:

```{r fit-spatial}
data("adjacency_matrix_car")
fit_car <- hbm_binlogitnorm(
  response  = "y",
  trials    = "n",
  auxiliary = c("x1", "x2", "x3"),
  sre       = "sre",                       # spatial-RE column
  sre_type  = "car",                       # CAR structure
  M         = adjacency_matrix_car,        # neighbour weight matrix
  data      = data_binlogitnorm,
  chains = 4, iter = 4000, warmup = 2000, cores = 4,
  seed = 1
)
summary(fit_car)
```

```
Correlation Structures:
      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sdcar     0.19      0.07     0.06     0.34 1.00     1843     2516

Regression Coefficients:
          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
Intercept     0.07      0.04     0.00     0.15 1.00     6018     6342
x1            0.26      0.03     0.20     0.32 1.00     7218     6917
x2           -0.14      0.03    -0.20    -0.08 1.00     6814     6539
x3            0.07      0.03     0.01     0.13 1.00     7104     6428
```

The new `Correlation Structures:` block reports the spatial standard
deviation `sdcar`.  See `vignette("hbsaems-spatial")` for SAR, BYM2, and
weight-matrix construction.

## Custom prior on a coefficient

To impose a tighter prior on a particular coefficient -- for example,
constraining the effect of an obviously non-informative auxiliary to
near-zero -- supply a `brms::set_prior()` object:

```{r fit-prior}
fit_prior <- hbm_binlogitnorm(
  response  = "y",
  trials    = "n",
  auxiliary = c("x1", "x2", "x3"),
  group     = "group",
  data      = data_binlogitnorm,
  prior     = brms::set_prior("normal(0, 0.1)", class = "b", coef = "x3"),
  chains = 4, iter = 4000, warmup = 2000, cores = 4,
  seed = 1
)
```

Default priors (`Intercept ~ student_t(4, 0, 10)`,
`b ~ student_t(4, 0, 2.5)`) are filled in for any class you don't
override.

## Missing data

If `y`, `n`, or auxiliary columns contain `NA`, the wrapper
**auto-selects** a strategy unless you supply `handle_missing`:

```{r fit-missing}
data_with_na <- data_binlogitnorm
data_with_na$x1[c(3, 14, 27)] <- NA          # 3 areas with missing x1

fit_miss <- hbm_binlogitnorm(
  response  = "y",
  trials    = "n",
  auxiliary = c("x1", "x2", "x3"),
  group     = "group",
  data      = data_with_na,
  # handle_missing not given -> auto-select "multiple" (mice)
  chains = 4, iter = 4000, warmup = 2000, cores = 4,
  seed = 1
)
```

```
Missing values detected. Auto-selecting handle_missing = 'multiple'
(family 'binomial', supports_mi = FALSE).

Missing predictor variable(s): x1. Applying multiple imputation
(mice) with m = 5 imputations.
Fitting imputed model 1...
Fitting imputed model 2...
...
Fitting imputed model 5...
```

Posterior samples from the five imputed fits are pooled via Rubin's
rules in the returned `hbmfit` object.  See
`vignette("hbsaems-handle-missing")` for the full discussion.

## Convergence and prediction

The diagnostic and prediction steps are the same as for any wrapper:

```{r conv-predict}
convergence_check(fit)

new_areas <- data.frame(
  group = 101:105,
  n     = c(150, 120, 180, 110, 140),
  x1    = rnorm(5),
  x2    = rnorm(5),
  x3    = rnorm(5)
)
preds <- sae_predict(fit, newdata = new_areas, allow_new_levels = TRUE)
preds
```

```
$verdict
[1] "All chains converged: Rhat < 1.01, ESS > 400, no divergent transitions."

  group estimate  lower  upper
1   101   0.524  0.461  0.589
2   102   0.461  0.395  0.531
3   103   0.587  0.523  0.648
4   104   0.448  0.378  0.519
5   105   0.495  0.432  0.560
```

Predictions for unseen areas use the marginal posterior of the random
intercept.

## Summary

* Use **`hbm_binlogitnorm()`** when you have *raw success counts* and
  *known trial counts*.
* Use **`hbm_betalogitnorm()`** instead when you have *direct
  proportion estimates with a design effect*.
* Add spatial structure via `sre`, `sre_type`, and `M` -- see the
  spatial vignette.
* Missing data is auto-handled with multiple imputation (mice).

## References

* Rao, J. N. K., & Molina, I. (2015). *Small Area Estimation*, 2nd ed.
  Wiley.  Chapter 4 covers area-level binary SAE.
* Buerkner, P.-C. (2017). brms: An R Package for Bayesian Multilevel
  Models Using Stan.  *Journal of Statistical Software*, 80(1), 1-28.
