---
title: "hbsaems-spatial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hbsaems-spatial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hbsaems)
```

# Introduction

In spatial data analysis, spatial effects play an important role in understanding patterns and relationships between locations. Spatial effects arise when the value of a variable in one location is affected by the value of the same variable in another location. This can happen due to natural processes (e.g. the spread of air pollution) or socio-economic phenomena (e.g. house prices in one area affect prices in surrounding areas).

This vignette aims to show how this package can be used to add spatial effects in Small Area Estimation Hierarchical Bayesian (SAEHB) modeling.

# Model Overview

The `hbsaems` package can handle spatial dependence effects (Spatial Autocorrelation) by using the :

1.  CAR (Conditional Autoregressive)
2.  SAR (Simultaneous Autoregressive)

## 1. CAR (Conditional Autoregressive)

Conditional Autoregressive (CAR) model is a spatial model that describes the relationship between adjacent areas through a local condition approach. In this model, the value of a variable in a location is directly influenced by the value of the same variable in the surrounding areas. CAR uses adjacency matrix to determine the relationship between locations.

### Simulated Data Example

```{r}
library(mlbench)

data("BostonHousing")
data_car <- BostonHousing

print(data_car)
```

To apply spatial effects to the data, we will cluster into 10 groups based on the variable `dis`. The variable `dis` in the `BostonHousing` dataset is the weighted distance to five job centers in Boston. This is a measure of accessibility to job centers. Clustering based on distance to the city center can reflect spatial patterns in the data so the variable `dis` can be a good proxy for creating spatial clustering.

```{r}
# Create a spatial grouping factor based on "dis"
data_car$spatial <- cut(data_car$dis, breaks = 10, labels = FALSE)  

library(Matrix)

data_car$spatial <- factor(data_car$spatial)
n <- length(levels(data_car$spatial))
M <- bandSparse(n = n, k = c(-1, 0, 1), 
                diag = list(rep(1, n - 1), rep(0, n), rep(1, n - 1)))
rownames(M) <- colnames(M) <- levels(data_car$spatial)
print(M)
```
In the spatial adjacency matrix, a value of 1 indicates that two spatial groups are neighbors based on their proximity to the city center, while a value of 0 indicates no adjacency relationship.

### Fitting the Model

We will try to model housing prices in each housing block using the Boston Housing dataset, where the target variable is the median owner-occupied home value (medv).

```{r}
model_car <- hbm(
  formula = bf(medv ~ crim + indus + rm + dis + rad + tax),  # Formula model
  hb_sampling = "gaussian",      # Gaussian family for continuous outcomes
  hb_link = "identity",          # Identity link function (no transformation)
  data = data_car,                   # Dataset
  sre = "spatial",                # Spatial random effect variable
  sre_type = "car",
  car_type = "icar",
  M = M,
  chains = 4,                    # Number of MCMC chains
  iter = 4000,                   # Total MCMC iterations
  warmup = 2000,                 # Number of warmup iterations
  cores = 2                      # Paralelisasi
)

summary(model_car)
```
When the `car_type` value is not specifically defined, its defaults to `escar`.

If `sre` is not specified, each observation is treated as a separate location. Therefore, in this case the dimension of the adjacency matrix of locations should be equal to the number of observations. It is recommended to always specify a grouping factor to allow for handling of new data in postprocessing methods.

## 2. SAR (Simultaneous Autoregressive)

The Simultaneous Autoregressive Model (SAR) handles spatial effects by defining dependence simultaneously across the study area. SAR uses a spatial weight matrix to describe the relationship between regions and can be applied in various forms, such as Spatial Lag Model (SLM) or Spatial Error Model (SEM), depending on how spatial effects affect the dependent variable or error in the model. `M` can be either the spatial weight matrix itself or an object of class listw or nb, from which the spatial weighting matrix can be computed.

The **lagsar** structure implements SAR of the response values:

$$y = \rho W y + \eta + e$$

The **errorsar** structure implements SAR of the residuals:

$$y = \eta + u, \quad u = \rho W u + e$$

In the above equations, $$\eta$$ is the predictor term and $$e$$ are independent normally or t-distributed residuals. Currently, only families **gaussian** and **student** support SAR structures.

### Simulated Data Example

```{r}
library(spdep)

data("oldcol")
data_sar <- COL.OLD

print(data_sar)
```


### Fitting the Model

```{r}
model_sar <- hbm(
  formula = bf(CRIME ~ INC + HOVAL),  
  hb_sampling = "gaussian",      # Gaussian family for continuous outcomes
  hb_link = "identity",          # Identity link function (no transformation)
  data = data_sar,                   # Dataset
  sre_type = "sar",
  car_type = "lag",
  M = COL.nb,
  chains = 2,                    # Number of MCMC chains
  iter = 4000,                   # Total MCMC iterations
  warmup = 2000,                 # Number of warmup iterations
  cores = 2                      # Paralelisasi
)

summary(model_sar)
```

When the `sar_type` value is not specifically defined, its defaults to `lag`.