---
title: "hbsaems-lognormal-model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hbsaems-lognormal-model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette demonstrates the usage of the `hbm_lognormal()` function from the `hbsaems` package for **Hierarchical Bayesian Small Area Estimation (HBSAE)** under a **Lognormal distribution**.

This method is particularly useful for modeling **small area estimates** when the response variable follows a lognormal distribution ($X>0$). The response variable is assumed to follow a lognormal distribution:

$$\log(y) \sim N(\mu, \sigma^2)$$

where $\mu$ represents the mean structure and $\sigma$ is the standard deviation in log-space. Internally, the model applies a log transformation to fit a normal model

## Load Required Packages

```{r setup}
library(hbsaems)
library(brms)
library(ggplot2)
```

## Simulated Data Example

Let's start by simulating data to fit a lognormal model.

```{r}
data("midwest")
data_midwest <- midwest[order(midwest$poptotal), ][2:31, ]
data_midwest$popbelowpoverty <- round(data_midwest$percbelowpoverty/100 * data_midwest$poptotal, 0)
head(data_midwest)
```

## Fitting the Model

We estimate the hierarchical Bayesian model using `hbm_lognormal()`

### 1. Basic Model

If a random effect (`re`) is not specified, the function automatically assigns `group` as a default:

```{r}
model <- hbm_lognormal(
  response = "popbelowpoverty",
  predictors = c("popdensity", "perchsd"),
  data = data_midwest
)
```

### 2. Model With Missing Data

```{r}
# Prepare Missing Data
data_midwest_missing <- data_midwest
data_midwest_missing$popbelowpoverty[sample(1:30, 3)] <- NA  # 3 missing values in response
```

-   **Handling missing data by deleted (Only if missing in response)**

    ```{r}
    model_deleted <- hbm_lognormal(
      response = "popbelowpoverty",
      predictors = c("popdensity", "perchsd"),
      data = data_midwest_missing,
      handle_missing = "deleted"
    )
    ```

-   **Handling missing data using model-based approach**

    ```{r}
    model_model <- hbm_lognormal(
      response = "popbelowpoverty",
      predictors = c("popdensity", "perchsd"),
      data = data_midwest_missing,
      handle_missing = "model"
    )
    ```

-   **Handling missing data using multiple imputation (m=5)**

    ```{r}
    model_multiple <- hbm_lognormal(
      response = "popbelowpoverty",
      predictors = c("popdensity", "perchsd"),
      data = data_midwest_missing,
      handle_missing = "multiple",
      m = 5
    )
    ```

### 3. Model With Random Effect

```{r}
model_re <- hbm_lognormal(
  response = "popbelowpoverty",
  predictors = c("popdensity", "perchsd"),
  re = "category",
  data = data_midwest
)
```

### 4. Model With Spatial Effect

```{r}
# Dummy spatial weight matrix
M <- matrix(0, 4, 4)
for (i in 1:(4 - 1)) {
  M[i, i + 1] <- 1
  M[i + 1, i] <- 1
}
# Set row and column names to match the group levels
rownames(M) <- unique(data_midwest$state)
colnames(M) <- unique(data_midwest$state)
```

```{r}
model_spatial <- hbm_lognormal(
  response = "popbelowpoverty",
  predictors = c("popdensity", "perchsd"),
  sre = "state",
  sre_type = "car",
  car_type = "esicar",
  M = M,
  data = data_midwest
)
```

## Model Summary

Inspect the model's summary to interpret parameter estimates and convergence diagnostics.

```{r}
summary(model)
```

## Model Diagnostics

We can assess model diagnostics using the hbcc function.

```{r}
diag_results <- hbcc(model)

# Rhat and ESS diagnostics
diag_results$rhat_ess

# Trace and density plots
diag_results$plots$trace
diag_results$plots$dens
```

## Model Comparison

If you have multiple models to compare, you can use the hbmc function. Here, we compare model with random effect and model with spatial random effect.

```{r}
comparison <- hbmc(model_re, model2 = model_spatial)
comparison$loo1_values
comparison$loo2_values
```

## Small Area Estimation Predictions

You can generate small area predictions using the hbsae function.

```{r}
sae_results <- hbsae(model)

# Prediction table
head(sae_results$pred_table)

# Mean Square Error (MSE) and Root Mean Square Error (RMSE)
sae_results$mse
sae_results$rmse
```

To generate predictions for new data:

```{r}
new_data <- midwest[31, c("popdensity", "perchsd")]
hbsae(model, newdata = new_data)
```

## Conclusion

The `hbm_lognormal()` function in **hbsaems** package provides a flexible way to fit a lognormal distribution. It incorporates random effect, spatial modeling, and handles missing data.This vignette has demonstrated the use of the hbm_lognormal function along with supporting functions (hbcc, hbmc, hbsae) for fitting, diagnosing, comparing, and predicting using a lognormal model for small area estimation.
