---
title: "Model Beta Logit Normal Small Area Estimation in hbsaems Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hbsaems-betalogitnorm-model}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

The `hbm_betalogitnorm` function in the `hbsaems` package is used for **Hierarchical Bayesian Small Area Estimation (HBSAE)** under the **Beta distribution**.

This method is particularly useful for modeling **small area estimates** when the response variable follows a beta distribution, allowing for efficient estimation of proportions or rates bounded between 0 and 1 while accounting for the inherent heteroskedasticity and properly modeling mean-dependent variance structures.

This vignette explains the function's usage including the handling of missing values and incorporating random effects (including spatial effects).

## Sample Model

$$
p_{iw} | P_i \sim \text{Beta}(a_i, b_i)
$$

with parameters :

$$
a_i = P_i \left( \frac{n_i}{\text{deff}_{iw}} - 1 \right) \\
b_i = \left( 1- P_i \right) \left( \frac{n_i}{\text{deff}_{iw}} - 1 \right)
$$
<<<<<<< HEAD
Where $p_{iw}$ is the proportion estimate for a small area of the survey (direct observation) and $P_i$ is the unknown value of the proportion parameter for a small area. $a_i$ and $b_i$ are the shape parameter of the Beta distribution. $n_i$ is the sample size for the area, while $\text{deff}$ is the Design Effect which corrects for the fact that the survey design is not simply random.
=======

Where $p_{iw}$ is the proportion estimate for a small area of the survey (direct observation) and $P_i$ is the unknown value of the proportion parameter for a small area.
$a_i$ and $b_i$ are the shape parameter of the Beta distribution.
$n_i$ is the sample size for the area, while $\text{deff}$ is the Design Effect which corrects for the fact that the survey design is not simply random.
>>>>>>> f97475e9fba6af00b5627f60fafb938a094de256

## Linking Model

The logit transformation is applied:

$$
\text{logit}(P_i) \mid \boldsymbol{\beta}, \sigma_\nu^2 \overset{\text{ind}}{\sim} \mathcal{N}(\mathbf{z}_i^T \boldsymbol{\beta}, \sigma_\nu^2)
$$

$\text{logit}(P_i)$ is logit transformation of proportion parameters, that is $\log \left( \frac{P_i}{1 - P_i} \right)$

# Simulated Data Example

We will use `data_betalogitnorm` to simulate the `hbm_betalogitnorm` function.
The `data_betalogitnorm` is a simulation data created specifically to demonstrate the implementation of *Hierarchical Bayesian Small Area Estimation* (HB SAE) with Beta distribution.
This data is suitable for testing Beta regression models with a hierarchical structure between areas.
This data is also equipped with variables that apply spatial effects.

This data consisting of 100 rows and 9 variables.

-   The `y` variable is the response variable in the form of the proportion of simulation results that have a value between 0 and 1 and follow the Beta distribution.

-   The `theta` variable is a latent parameter related to the mean of the Beta distribution.

-   Three predictor variables, namely `x1`, `x2`, and `x3`, are used to model variations in `y`.

-   The `n` variable indicates the number of sample units in each area or region used in the survey - `deff` states the design effect of the survey.

-   The `group` variable is the area ID (1–100) used in random effects modeling to capture heterogeneity between regions.

-   In addition, there is a `sre` variable which is an optional grouping factor to map observations to a specific spatial location.

-   To apply spatial effects, we will also use the `adjacency_matrix_car` data available in this package.

```{r}
library(hbsaems)
data("data_betalogitnorm")
data <- data_betalogitnorm

data("adjacency_matrix_car")
M <- adjacency_matrix_car
```

# Initial Model

We begin by specifying the model with informative priors on the coefficients and the intercept.
The `sample_prior = "only"` argument ensures that the model ignores the data and simulates predictions based on the priors alone.
This is particularly useful for performing a **prior predictive check**, which involves generating data purely from the prior distributions to evaluate whether the priors lead to plausible values of the outcome variable.

```{r}
model_prior_pred <- hbm_betalogitnorm(
   response = "y",
   predictors = c("x1", "x2", "x3"),
   data = data,
   sample_prior = "only",
   prior = c(
    set_prior("normal(-1, 0.7)", class = "Intercept"),
    set_prior("normal(0, 0.5)", class = "b"),
    set_prior("gamma(2, 0.1)", class = "phi")
  )
)
```

```{r}
summary(model_prior_pred)
```

If you have information about the values of `n` and `deff` then you do not need to provide prior information for phi because the value of phi is obtained from the calculation $\phi = \left( \frac{n_i}{\text{deff}_{iw}} - 1 \right)$.

```{r}
model_prior_pred_phi <- hbm_betalogitnorm(
   response = "y",
   predictors = c("x1", "x2", "x3"),
   n = "n",
   deff = "deff",
   data = data,
   sample_prior = "only",
   prior = c(
    set_prior("normal(-1, 0.7)", class = "Intercept"),
    set_prior("normal(0, 0.5)", class = "b")
  )
)
```

```{r}
summary(model_prior_pred_phi)
```

-   The intercept prior, **normal(-1, 0.7)**, reflects our belief that on the logit scale, the baseline probability (when all predictors are zero) is centered at **approximately 27%** (since $logit^{-1}(-1) ≈ 0.27$).
    This prior places most of its mass roughly between **10% and 53%** (±1 standard deviation: $logit^{-1}(-1.7) ≈ 0.15$, $logit^{-1}(-0.3) ≈ 0.43$).
    Such a prior avoids extreme baseline probabilities near 0 or 1, helping ensure that prior predictive distributions remain within a realistic and interpretable range.

-   The coefficient prior for the predictors, **normal(0, 0.5)**, constrains the effect sizes to be modest.
    On the odds ratio scale, this implies that for each unit change in a predictor, the odds are expected to change by a factor between **\~0.61 and \~1.65** ($exp(-0.5) ≈ 0.61$, $exp(0.5) ≈ 1.65$ ) with 68% probability.
    This prevents the model from assigning overly large predictor effects that could push predicted probabilities too close to the boundaries (0 or 1).

# Prior Predictive Check

Before fitting a Bayesian model to the data, it is important to evaluate whether the chosen priors lead to sensible predictions.
This process is called a **prior predictive check**.

Prior predictive checks help to:

-   Identify whether the priors are overly vague or unrealistically informative.
-   Ensure that the prior assumptions do not produce unreasonable predictions before any actual data are used.
-   Avoid unintentionally strong or misleading prior beliefs that could affect posterior inference.

This step is especially important in Bayesian modeling to build confidence that the model structure and prior choices are coherent before observing any data.

```{r}
result_hbpc <- hbpc(model_prior_pred, response_var="y")
summary(result_hbpc)
```

```{r}
result_hbpc$prior_predictive_plot
```

The prior predictive plot indicates that the priors used in the model are reasonable and appropriate.
Here's why:

-   The **prior predictive distributions** (shown in blue) cover a wide but plausible range of values for the response variable, which aligns well with the distribution of the **observed data** (shown in black).

-   The shape of the simulated predictions is **consistent with the general pattern** of the observed outcome, without being overly concentrated or too diffuse.

-   There are **no extreme or unrealistic values** produced by the priors, suggesting that the priors are informative enough to guide the model without being too restrictive.

# Fit a Beta Model

After prior predictive checks confirm the priors are suitable, we proceed to fit the model using `sample_prior = "no"`.
Why `sample_prior = "no"`?

-   We already conducted a prior predictive check (previous explanation), `sample_prior = "no"` tells `brms` to:

    -   Use the **priors** in the estimation.

    -   Focus entirely on drawing from the **posterior** given the observed data.

-   This option is standard for final model fitting after priors have been validated.

**From this stage to the next will be explained the construction of the model with the condition that the user has information on the value of n and deff. If you do not have information related to the value of n and deff then simply delete the parameters n and deff in your model.**

## 1. Basic Model

```{r}
model <- hbm_betalogitnorm(
   response = "y",
   predictors = c("x1", "x2", "x3"),
   n = "n",
   deff = "deff",
   data = data,
   sample_prior = "no",
   prior = c(
     set_prior("normal(-1, 0.7)", class = "Intercept"),   
     set_prior("normal(0, 0.5)", class = "b")
  )
)
```

```{r}
summary(model)
```

By default, if we do not explicitly define a random effect, the model will still generate one based on the natural random variations between individual records (rows) in the dataset.
However, we can also explicitly define a random effect to account for variations at a specific hierarchical level, such as neighborhoods or residential blocks.
We can use parameter group.

```{r}
model_with_defined_re <- hbm_betalogitnorm(
   response = "y",
   predictors = c("x1", "x2", "x3"),
   n = "n",
   deff = "deff",
   group = "group",
   data = data,
   sample_prior = "no",
   prior = c(
     set_prior("normal(-1, 0.7)", class = "Intercept"),   
     set_prior("normal(0, 0.5)", class = "b")
  )
)
```

```{r}
summary(model_with_defined_re)
```

## 2. Model With Missing Data

The hbm function supports three strategies for handling missing data:

1.  "deleted": Removes rows with missing values.

2.  "multiple": Performs multiple imputation using mice.

3.  "model": Uses the mi() function to model missingness directly (not available for discrete outcomes).

-   **Handling missing data by deleted (Only if missing in response)**

    ```{r}
    # Prepare Missing Data
    data_missing <- data
    data_missing$y[sample(1:30, 3)] <- NA  # 3 missing values in response
    ```

    ```{r}
    model_deleted <- hbm_betalogitnorm(
       response = "y",
       predictors = c("x1", "x2", "x3"),
       n = "n",
       deff = "deff",
       group = "group",
       data = data_missing,
       handle_missing = "deleted",
       sample_prior = "no",
       prior = c(
         set_prior("normal(-1, 0.7)", class = "Intercept"),   
         set_prior("normal(0, 0.5)", class = "b")
      )
    )
    ```

    ```{r}
    summary(model_deleted)
    ```

-   **Handling missing data** **before model fitting using multiple imputation**

    ```{r}
    # Prepare missing data
    data_missing <- data
    data_missing$y[3:5] <- NA 
    data_missing$x1[6:7] <- NA
    ```

    ```{r}
    model_multiple <- hbm_betalogitnorm(
       response = "y",
       predictors = c("x1", "x2", "x3"),
       n = "n",
       deff = "deff",
       group = "group",
       data = data_missing,
       handle_missing = "multiple",
       sample_prior = "no",
       prior = c(
         set_prior("normal(-1, 0.7)", class = "Intercept"),   
         set_prior("normal(0, 0.5)", class = "b")
      )
    )
    ```

    ```{r}
    summary(model_multiple)
    ```

-   **Handle missing data during model fitting using mi()**

    If we want to handle missing data in the model, we must define the formula correctly.
    In this case, we need to specify missing data indicators using the `mi()` function.

    ```{r}
    data_missing <- data
    data_missing$x1[3:5] <- NA 
    data_missing$x2[14:17] <- NA 
    ```

    ```{r}
    model_during_model <- hbm_betalogitnorm(
       response = "y",
       predictors = c("x1", "x2", "x3"),
       n = "n",
       deff = "deff",
       group = "group",
       data = data_missing,
       handle_missing = "model",
       sample_prior = "no",
       prior = c(
         set_prior("normal(-1, 0.7)", class = "Intercept"),   
         set_prior("normal(0, 0.5)", class = "b")
      )
    )
    ```

    ```{r}
    summary(model_during_model)
    ```

## 3. Model With Spatial Effect

For the Beta distribution, this package supports only the Conditional Autoregressive (CAR) spatial structure.
The Spatial Autoregressive (SAR) model is currently available only for Gaussian and Student's t families.
A more detailed explanation of the use of spatial effects can be seen in the spatial vignette in this package.

```{r}
model_spatial <- hbm_betalogitnorm(
   response = "y",
   predictors = c("x1", "x2", "x3"),
   n = "n",
   deff = "deff",
   group = "group",
   sre = "sre",
   sre_type = "car",
   car_type = "icar",
   M = M,
   data = data,
   sample_prior = "no",
   prior = c(
     set_prior("normal(-1, 0.7)", class = "Intercept"),   
     set_prior("normal(0, 0.5)", class = "b")
  )
)
```

```{r}
summary(model_spatial)
```

# Model Diagnostics

The `hbcc` function is designed to evaluate the convergence and quality of a Bayesian hierarchical model.
It performs several diagnostic tests and generates various plots to assess Markov Chain Monte Carlo (MCMC) performance.

```{r}
result_hbcc <- hbcc(model)
summary(result_hbcc)
```

The `update_hbm()` function allows you to continue sampling from an existing fitted model by increasing the number of iterations.
This is particularly useful when initial model fitting did not achieve convergence, for example due to insufficient iterations or complex posterior geometry.

When `update_hbm()` is called with additional iterations, the sampler resumes from the previous fit, effectively increasing the total number of posterior samples.
This helps improve convergence diagnostics such as Rhat, effective sample size, and chain mixing.

```{r}
model <- update_hbm(model, iter = 8000)
```

```{r}
summary(model)
```

```{r}
result_hbcc <- hbcc(model)
summary(result_hbcc)
```

The **trace plot** (`result_hbcc$plots$trace`) shows the sampled values of each parameter across MCMC iterations for all chains.
A well-mixed and stationary trace (with overlapping chains) indicates good convergence and suggests that the sampler has thoroughly explored the posterior distribution.

```{r}
result_hbcc$plots$trace
```

The **density plot** (`result_hbcc$plots$dens`) displays the estimated posterior distributions of the model parameters.
When the distributions from multiple chains align closely, it supports the conclusion that the chains have converged to the same target distribution.

```{r}
result_hbcc$plots$dens
```

The **autocorrelation plot (ACF)** (`result_hbcc$plots$acf`) visualizes the correlation between samples at different lags.
High autocorrelation can indicate inefficient sampling, while rapid decay in autocorrelation across lags suggests that the chains are generating nearly independent samples.

```{r}
result_hbcc$plots$acf
```

The **NUTS energy plot** (`result_hbcc$plots$nuts_energy`) is specific to models fitted using the No-U-Turn Sampler (NUTS).
This plot examines the distribution of the Hamiltonian energy and its changes between iterations.
A well-behaved energy plot indicates stable dynamics and efficient exploration of the parameter space.

```{r}
result_hbcc$plots$nuts_energy
```

The **Rhat plot** (`result_hbcc$plots$rhat`) presents the potential scale reduction factor (R̂) for each parameter.
R̂ values close to 1.00 (typically \< 1.01) are a strong indicator of convergence, showing that the between-chain and within-chain variances are consistent.

```{r}
result_hbcc$plots$rhat
```

The **effective sample size (n_eff) plot** (`result_hbcc$plots$neff`) reports how many effectively independent samples have been drawn for each parameter.
Higher effective sample sizes are desirable, as they indicate that the posterior estimates are based on a substantial amount of independent information, even if the chains themselves are autocorrelated.

```{r}
result_hbcc$plots$neff
```

# Model Comparison

The `hbmc` function is used to compare Bayesian hierarchical models and assess their posterior predictive performance.
It allows for model comparison, posterior predictive checks, and additional diagnostic visualizations.

```{r}
result_hbmc <- hbmc(
      model = list(model, model_spatial),
      comparison_metrics = c("loo", "waic", "bf"),
      run_prior_sensitivity= TRUE, 
      sensitivity_vars = c ("b_x1")
  )
  
summary(result_hbmc)
```

The **posterior predictive check plot** (`result_hbmc$primary_model_diagnostics$pp_check_plot`) compares the observed data to replicated datasets generated from the posterior predictive distribution.
This visualization helps evaluate how well the model reproduces the observed data.
A good model fit is indicated when the simulated data closely resemble the actual data in distribution and structure.

```{r}
result_hbmc$primary_model_diagnostics$pp_check_plot
```

The **marginal posterior distributions plot** (`result_hbmc$primary_model_diagnostics$params_plot`) displays the estimated distributions of the model parameters based on the posterior samples.
This plot is useful for interpreting the uncertainty and central tendency of each parameter.
Peaks in the distribution indicate likely values, while wider distributions suggest greater uncertainty.

```{r}
result_hbmc$primary_model_diagnostics$params_plot
```

# Small Area Estimation Predictions

The `hbsae()` function implements Hierarchical Bayesian Small Area Estimation (**HBSAE**) using the `brms` package.
It generates small area estimates while accounting for uncertainty and hierarchical structure in the data.
The function calculates Mean Predictions, Relative Standard Error (RSE), Mean Squared Error (MSE), and Root Mean Squared Error (RMSE) based on the posterior predictive sample from the fitted Bayesian model.

```{r}
result_hbsae <- hbsae(model)
summary(result_hbsae)
```

# Conclusion

The `hbm_betalogitnorm` function in the hbsaems package provides a flexible framework for modeling proportions or rates using the Beta distribution, particularly suited for small-area estimation where the response lies within the (0, 1) interval.
This function supports random effects and spatial modeling through CAR structures, enhancing estimation in the presence of spatial autocorrelation.
It also accommodates missing data through deletion or multiple imputation approaches.
Diagnostic tools such as convergence assessment and posterior predictive checks ensure model adequacy, while model comparison features allow users to assess the role of spatial effects or alternative specifications.
Predictions are derived from the posterior distributions, offering robust uncertainty quantification.
This vignette has demonstrated the use of `hbm_betalogitnorm` along with supporting functions (`hbcc`, `hbmc`, `hbsae`) for fitting, diagnosing, comparing, and predicting with a Beta model in small-area estimation.
