---
title: "Interactive Modelling with the SAE Shiny Application"
output: rmarkdown::html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

```{r preload, eval = TRUE, include = FALSE}
suppressPackageStartupMessages(library(hbsaems))
```

> **Note.**  This vignette describes the interactive **Shiny
> application** bundled with `hbsaems`.  Because Shiny apps are
> stateful and interactive, the screenshots and example workflows
> below are **descriptive** -- they show what the user sees when
> running the app locally.  Stan fits triggered from within the GUI
> are not run at vignette-build time.

## When to use the Shiny app

The bundled Shiny application, launched by `run_sae_app()`, is intended
for two audiences:

* **Non-programmers**: subject-matter experts, BPS staff, NGO
  analysts, or students who want to fit and inspect an SAE model
  without writing R code.
* **Programmers prototyping a model**: a quick way to explore the
  effect of priors, link functions, or spatial structure on a small
  dataset before committing to a scripted workflow.

The app is **not** meant to replace scripted modelling for production
work -- the audit trail of an `.R` script is essential there.

## Launching the app

```{r launch}
library(hbsaems)
run_sae_app()
```

This launches the app in your default browser.  By default it runs
on a random local port; pass `port = 4242` to fix the port.

The app calls `check_shiny_deps()` at startup; if the required Shiny
packages are missing it prints an actionable install message:

```
The Shiny application requires the following packages:
  - shiny (>= 1.7.0)
  - shinydashboard
  - DT
  - bslib

Install them with:
  install.packages(c("shiny", "shinydashboard", "DT", "bslib"))
```

## Layout overview

The application is organised into six sidebar sections:

```
+------------------------+-----------------------------------------------+
|  Sidebar               |  Main panel (sub-tabs)                        |
+========================+===============================================+
|  Data Upload           |  CSV / RDS / use built-in datasets            |
|  Spatial Setup         |  Source / Map / Matrix / Diagnostic / Ref     |
|  Model Specification   |  Family, formula, priors, RE / spatial RE     |
|  MCMC Settings         |  Chains, iter, warmup, cores, control         |
|  Fit & Diagnostics     |  Run, summary, traceplot, posterior, LOO       |
|  Predictions           |  Predict for unsampled areas, export CSV       |
+------------------------+-----------------------------------------------+
```

A language selector at the top of the sidebar toggles the UI between
English and Bahasa Indonesia (`tr()`-driven; see
`vignette("complete-workflow")` for the i18n architecture).

## Section 1: Data Upload

Three options:

1. **Upload a CSV / RDS file** -- standard file-picker; the app
   previews the first 10 rows and lets you mark columns as response,
   auxiliary, group, or trial/n.
2. **Use a bundled dataset** -- dropdown selector with `data_fhnorm`,
   `data_lnln`, `data_betalogitnorm`, `data_binlogitnorm`.
3. **Paste from clipboard** -- for quick experiments with copied
   spreadsheet data.

Every uploaded dataset is validated by `check_data()`, which reports:

* Total rows and columns
* Missing-value counts per column
* Response distribution check (e.g.\ "98% in (0, 1)" for proportions)

## Section 2: Spatial Setup

A dedicated section since spatial modelling is the trickiest part of
SAE for newcomers.  Five sub-tabs:

* **Source**: choose between (a) a shapefile / GeoJSON upload,
  (b) a precomputed adjacency matrix CSV, or (c) one of the bundled
  example matrices.
* **Map**: leaflet-based preview of the polygons coloured by the
  response variable.
* **Matrix**: heatmap of the weight matrix, with row-normalisation
  and standardisation options.
* **Diagnostic**: runs `check_spatial_weight()` and reports
  symmetry, isolates, mean neighbours, and a verdict.
* **Reference**: short text on CAR, SAR, BYM2, and the recommended
  parameterisations.

## Section 3: Model Specification

Step-by-step:

1. **Likelihood family**: dropdown (`gaussian`, `lognormal`,
   `Beta`, `binomial`, `loglogistic`, `shifted_loglogistic`, ...).
   Custom families registered via `register_hbsae_brms_custom()`
   appear in the same list.
2. **Response variable**: dropdown of numeric columns; the app
   greys out incompatible options (e.g.\ negative values for a
   lognormal).
3. **Auxiliary variables**: multi-select.
4. **Group (random intercept)**: dropdown of integer/character
   columns.
5. **Spatial random effect**: toggle on/off; if on, choose CAR or
   SAR.
6. **Pinned parameters**: toggle on/off for $\phi$ from `n`+`deff`
   (Beta) or $\sigma$ from `sampling_variance` (lnln).
7. **Priors**: a table of priors with reasonable defaults that the
   user can edit.  Distribution preview (PDF/CDF plots) is shown
   for the currently-selected class.

The "Build formula" button assembles the `bf(...)` call and shows
the equivalent R code in a code block, which can be copied for
scripted use:

```r
# Equivalent R code (copy this for a scripted workflow):
hbm_lnln(
  response     = "y_obs",
  auxiliary    = c("x1", "x2", "x3"),
  area_var   = "regency",
  sampling_variance = "psi_i",
  data         = my_data,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

## Section 4: MCMC Settings

Sliders for `chains` (1-8), `iter` (1000-10000), `warmup`
(fraction of iter), `cores` (capped at `parallel::detectCores()`),
and `seed`.  Advanced control options (`adapt_delta`, `max_treedepth`)
are exposed under an "Advanced" disclosure.

## Section 5: Fit & Diagnostics

Once "Fit Model" is clicked, the app:

1. Disables the Fit button and shows a progress bar with the live
   Stan output (compilation + each chain's iteration count).
2. After the fit completes, populates four diagnostic sub-tabs:
   * **Summary**: the standard `summary()` output rendered as an
     HTML table.
   * **Traceplot**: `bayesplot::mcmc_trace()` for selected
     parameters.
   * **Posterior**: density plots / `mcmc_areas` for selected
     parameters; "Compare with prior" toggle.
   * **LOO**: `loo()` output and Pareto-$k$ diagnostic plot.

Saving the fit creates an `.rds` file with the `hbmfit` object.

## Section 6: Predictions

Two modes:

* **In-sample**: posterior mean and 95% CI for every area in the
  original dataset, joined with the area ID, exported as CSV.
* **Out-of-sample**: upload (or paste) auxiliary values for new
  areas; the app calls `sae_predict(..., allow_new_levels = TRUE)`
  and produces a downloadable predictions table plus, if a polygon
  layer is loaded, a choropleth map.

## Reproducibility

Every action in the app is logged to a hidden R log that can be
exported as a runnable `.R` script.  This gives users the best of
both worlds: interactive exploration plus a scripted record that
reproduces the entire session.

## Limitations

* **Single-user**.  The app is designed to be run locally
  (`run_sae_app()` on your laptop); it is not a multi-user web
  service.  For a deployable web service, use Shiny Server or
  Posit Connect with the bundled app source at
  `system.file("shiny", "sae_app", package = "hbsaems")`.
* **Dataset size**.  In-memory only; not designed for datasets
  larger than a few thousand areas.
* **Custom families**.  Custom brms families must be registered
  *before* launching the app (`register_hbsae_brms_custom()`); they
  cannot be defined from inside the app.
* **Stan compilation time** is unaffected by the GUI; the first fit
  in a session still incurs the one-time compile cost.

## See also

* `vignette("hbsaems-modelling")` -- the scripted workflow this app
  parallels.
* `?run_sae_app` -- function reference.
* `system.file("shiny", "sae_app", package = "hbsaems")` -- the app
  source, suitable for adaptation.
