---
title: "Advanced Features: Spatial, Benchmarking, Custom Families, Smooth Terms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Features: Spatial, Benchmarking, Custom Families, Smooth Terms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This vignette demonstrates the advanced features available in
**hbsaems v1.0.0**, complementing the
`complete-workflow` vignette which focuses on the core fitting pipeline.

Topics covered:

1.  **Pre-fit data quality checking** (`check_data`)
2.  **Spatial weight matrices** -- building from shapefiles + theory checks
3.  **Multi-domain benchmarking** -- kecamatan to kabupaten with raking
4.  **Fully Bayesian benchmarking** -- correct uncertainty after adjustment
5.  **Smooth terms** -- splines and Gaussian processes
6.  **Shrinkage priors** -- horseshoe and R2D2
7.  **Custom distributions** -- the model registry

```{r}
library(hbsaems)
```

---

## 1. Pre-fit data quality checking

Before fitting any model, run `check_data()` to verify the variables
exist, identify missing patterns, and get a recommendation for the
`handle_missing` strategy.

```{r}
data("data_fhnorm")

chk <- check_data(
  data       = data_fhnorm,
  response   = "y",
  predictors = c("x1", "x2", "x3"),
  sre        = "sre"
)

print(chk)
#>  HBSAE Data Check  [hbsaems_data_check]
#>  ---------------------------------------
#>   Observations    : 50
#>   Missing pattern : none (data complete)
#>   Issues          : none
#>
#>   - No missing values detected. handle_missing can stay NULL.
```

`check_data()` returns an object of class
`c("hbsaems_data_check", "hbsaems_check")`, so you can test it with the
generic predicate:

```{r}
is.hbsaems_check(chk)   # TRUE
```

### When the response has missing values

Missing values in *Y* often correspond to **non-sample areas** -- domains
for which you want a prediction but have no direct estimate.
`check_data()` highlights this explicitly:

```{r}
d2 <- data_fhnorm
d2$y[1:5] <- NA
chk2 <- check_data(d2, response = "y", predictors = c("x1", "x2", "x3"))
chk2$non_sample_warning
```

If those rows really are non-sample areas, **keep them**: fit the model
on the complete-Y subset and use `sae_predict(..., newdata = na_rows)`.

---

## 2. Spatial weight matrices

### Building W from a shapefile

`build_spatial_weight()` constructs a CAR-ready binary matrix or a
SAR-ready row-standardised matrix from a polygon shapefile:

```{r}
# Queen contiguity + binary -> for CAR (Besag 1974)
M_car <- build_spatial_weight(
  shp       = "path/to/areas.shp",
  for_model = "car"         # implies type = "queen", style = "B"
)

# K-nearest + row-standardised -> for SAR (Anselin 1988)
M_sar <- build_spatial_weight(
  shp       = "path/to/areas.shp",
  for_model = "sar",        # implies type = "knn", style = "W"
  k         = 4
)
```

Distance-band and rook contiguity are also supported -- see
`?build_spatial_weight`.

### Theoretical compatibility check

`check_spatial_weight()` runs five checks against the requirements of
the chosen model class:

```{r}
check_spatial_weight(M_car, sre_type = "car")
#> Spatial Weight Matrix Diagnostic
#> ---------------------------------
#>   Square          : TRUE
#>   Zero diagonal   : TRUE
#>   Symmetric       : TRUE
#>   Detected style  : B
#>   Isolated areas  : 0
#>   Components      : 1
#>   Matrix is theoretically compatible.
```

This catches the common mistakes: non-symmetric matrices fed into CAR,
isolated areas under ICAR, etc.  It also runs automatically inside
`hbm()` whenever `sre_type` is set.

---

## 3. Multi-domain benchmarking: kecamatan to kabupaten

A common SAE scenario: you have estimates per kecamatan and need them
to roll up to a published kabupaten total from BPS.  `sae_benchmark()`
with `method = "raking"` handles this:

```{r}
# Suppose we have SAE estimates per kecamatan, with a kabupaten column
estimates <- sae_predict(fit_hbm)

# Official BPS totals
T_kab <- c(Bogor = 110, Sukabumi = 145, Cianjur = 145)

bm <- sae_benchmark(
  predictions = estimates,
  target      = T_kab,                  # one target per kabupaten
  weights     = data_fhnorm$populasi,   # population per kecamatan
  groups      = data_fhnorm$kabupaten,  # mapping kec -> kab
  method      = "raking"
)

bm$benchmark_info$converged    # raking with one group var: converges in 1 sweep
```

The adjustment factor is *constant within each kabupaten* but
*different across kabupaten* -- exactly what multi-domain benchmarking
requires.

### Unit consistency

The critical rule: `sum(weights * theta) == target`, with consistent
units.  Two common pitfalls:

| Wrong                              | Right                                  |
|------------------------------------|----------------------------------------|
| weights in raw count, target in %  | both as totals OR both as means        |
| theta in proportion, weights as N  | weights = N/total, then target as rate |

---

## 4. Fully Bayesian benchmarking (v1.0.0)

The default `sae_benchmark()` mode applies the adjustment to the
posterior **mean** only -- the SD column is approximated as unchanged.
**This understates uncertainty for multiplicative ratios** and over- or
understates it for additive shifts.

The fully Bayesian mode applies the benchmark to **every posterior
draw** and recomputes SD, RSE, and quantiles correctly:

```{r}
# Pass posterior = TRUE if predictions object carries draws
bm_bayes <- sae_benchmark(
  predictions = estimates,
  target      = 1000,
  method      = "ratio",
  posterior   = TRUE,                # NEW in v1.0.0
  probs       = c(0.025, 0.5, 0.975)
)

bm_bayes$result_table
#>    Prediction      SD  RSE_percent      Q025      Q50      Q975
#> 1     ...           ...     ...          ...      ...      ...

# Or supply your own draws matrix (D x n)
draws <- posterior_predict(fit$model)
bm_bayes2 <- sae_benchmark(estimates, target = 1000,
                            method = "ratio",
                            posterior = draws)
```

### Why this matters

For multiplicative ratio benchmarking, the adjustment factor itself is
a random variable computed from the same draws:
\[
  r_d = T \, / \, \sum_j w_j \, \theta_{d,j},
  \qquad
  \theta_{d,i}^{B} = \theta_{d,i} \cdot r_d.
\]
Since $\theta_{d,i}$ and $r_d$ are negatively correlated (a draw with
larger area-$i$ value implies a smaller $r_d$), the post-benchmark
variance is **strictly smaller** than the naive $r^2 \, \mathrm{Var}(\theta)$
approximation.  Fully Bayesian benchmarking captures this exactly.

---

## 5. Smooth terms: splines and Gaussian processes

Replace any linear predictor with a smooth function:

```{r}
# Penalised regression spline (mgcv backend)
fit_spline <- hbm(
  formula        = brms::bf(y ~ x1 + x2 + x3),
  data           = data_fhnorm,
  nonlinear      = c("x2", "x3"),     # variables to smooth
  nonlinear_type = "spline",
  spline_k       = 7L,                 # basis dimension (-1 = auto)
  chains = 2, iter = 2000, refresh = 0
)

# Gaussian process
fit_gp <- hbm(
  formula        = brms::bf(y ~ x1 + x2),
  data           = data_fhnorm,
  nonlinear      = c("x1"),
  nonlinear_type = "gp",
  gp_scale       = 1.5,                # length-scale 'c' (NULL -> brms default)
  chains = 2, iter = 2000, refresh = 0
)
```

The same arguments work in the distribution-specific wrappers
(`hbm_lnln`, `hbm_binlogitnorm`, `hbm_betalogitnorm`) and in
`hbm_flex()` (the flexible registry-aware factory).

---

## 6. Shrinkage priors

### Horseshoe (Piironen & Vehtari 2017)

```{r}
fit_hs <- hbm(
  formula        = brms::bf(y ~ x1 + x2 + x3 + x4 + x5),
  data           = my_data,
  prior_type     = "horseshoe",
  hs_df          = 1,        # local half-t df  (1 = original HS)
  hs_df_global   = 1,        # global half-t df
  hs_df_slab     = 4,        # slab half-t df   (regularised HS+)
  hs_scale_slab  = 2,        # slab scale
  chains = 2, iter = 2000, refresh = 0
)
```

### R2D2 (Zhang, Yang & Bondell 2022)

```{r}
fit_r2d2 <- hbm(
  formula        = brms::bf(y ~ x1 + x2 + x3 + x4 + x5),
  data           = my_data,
  prior_type     = "r2d2",
  r2d2_mean_R2   = 0.5,      # prior mean of R^2
  r2d2_prec_R2   = 2,        # prior precision of R^2
  chains = 2, iter = 2000, refresh = 0
)
```

Both priors are **cross-cutting** -- they work with any family
(Gaussian, Beta, binomial, lognormal, custom...) and combine with
splines / GPs without extra setup.

---

## 7. Custom distributions: the model registry

The model registry (`R/models-registry.R`) catalogues every model spec
hbsaems knows about.  Each spec bundles a likelihood family, a default
link, optional auxiliary-parameter hyperpriors, missing-data support
flags, and response-domain checks.  Registered models are usable via
`hbm()`, the convenience wrappers, and the flexible factory
`hbm_flex()`.

### Inspecting the registry

```{r}
list_hbsae_models()
#> [1] "bernoulli" "Beta" "beta-binomial" "binomial" ...

get_hbsae_model("beta")
#> $family             : "Beta"
#> $link               : "logit"
#> $discrete           : FALSE
#> $supports_mi        : TRUE
#> $aux_param_hyperprior: function(args, data) ...   # phi ~ gamma(alpha, beta)
```

### Adding a custom model (Tier-1: pure metadata)

For distributions that just need a name, link, and domain check:

```{r}
register_hbsae_model(
  key                = "gamma_log",
  family             = "Gamma",
  link               = "log",
  discrete           = FALSE,
  supports_mi        = TRUE,
  response_check     = function(y) all(y > 0, na.rm = TRUE),
  response_check_msg = "Gamma response must be strictly positive."
)

# Use immediately
fit_gamma <- hbm_flex(
  family_key = "gamma_log",
  response   = "expenditure",
  predictors = c("age", "income", "education"),
  data       = my_household_data
)
```

### Adding a custom model (Tier-2: auxiliary parameter hyperprior)

For distributions with auxiliary parameters (e.g.\ $\phi$ for Beta,
shape for Gamma):

```{r}
register_hbsae_model(
  key            = "gamma_with_hyperprior",
  family         = "Gamma",
  link           = "log",
  response_check = function(y) all(y > 0, na.rm = TRUE),

  # Inject Stan code via stanvars + a prior on the shape parameter
  aux_param_hyperprior = function(args, data) {
    if (!isTRUE(args$add_shape_prior)) return(NULL)
    list(
      stanvars = brms::stanvar(
                  scode = "real<lower=0> shape_a;\n  real<lower=0> shape_b;",
                  block = "parameters") +
                brms::stanvar(
                  scode = "shape_a ~ gamma(1, 1);\n  shape_b ~ gamma(1, 1);",
                  block = "model"),
      prior    = brms::set_prior("gamma(shape_a, shape_b)", class = "shape")
    )
  }
)

# Trigger the hyperprior via aux_args
fit <- hbm_flex(
  family_key = "gamma_with_hyperprior",
  response   = "y",
  predictors = c("x1", "x2"),
  data       = my_data,
  aux_args   = list(add_shape_prior = TRUE)
)
```

---

## Putting it all together

A realistic workflow combining everything:

```{r}
# 1. Check data
chk <- check_data(my_data, response = "y",
                   predictors = c("x1", "x2", "x3"),
                   sre = "kecamatan")

# 2. Build CAR matrix from shapefile
M <- build_spatial_weight("kecamatan.shp",
                            for_model = "car",
                            id_col    = "kec_code")

# 3. Configure with helpers (v1.0.0+); pass bundles directly to hbm()
priors <- hbm_priors(prior_type = "horseshoe", hs_df_slab = 4)
nl     <- hbm_nonlinear(c("x1"), type = "spline", k = 5)
ctrl   <- hbm_control(chains = 4, iter = 4000, cores = 4,
                       adapt_delta = 0.95)

# 4. Fit
fit <- hbm(
  formula = brms::bf(y ~ x1 + x2 + x3),
  data    = my_data,
  sre = "kecamatan", sre_type = "car", M = M,
  handle_missing = chk$recommended_method,
  priors, nl, ctrl
)

# 5. Predict at all areas (including non-sample)
est <- sae_predict(fit)

# 6. Benchmark to kabupaten totals (fully Bayesian)
T_kab <- c(Bogor = 110, Sukabumi = 145, Cianjur = 145)
bm <- sae_benchmark(est,
                     target  = T_kab,
                     weights = my_data$populasi,
                     groups  = my_data$kabupaten,
                     method  = "raking",
                     posterior = TRUE)

# 7. Use the corrected uncertainty
bm$result_table
```

---

## References

* **Besag, J.** (1974). Spatial interaction and the statistical analysis
  of lattice systems. *JRSS-B*, 36(2), 192-225.
* **Anselin, L.** (1988). *Spatial Econometrics: Methods and Models*.
  Kluwer.
* **Pfeffermann, D.** (2013). New important developments in small area
  estimation. *Statistical Science*, 28(1), 40-68.
* **Piironen, J. & Vehtari, A.** (2017). Sparsity information and
  regularization in the horseshoe and other shrinkage priors.
  *Electronic J. of Statistics*, 11(2), 5018-5051.
* **Zhang, Y. D., Naughton, B. P., Bondell, H. D., & Reich, B. J.**
  (2022). Bayesian Regression Using a Prior on the Model Fit: The R2-D2
  Shrinkage Prior. *JASA*, 117(538), 862-874.
