---
title: "Modelling Workflow Overview with hbsaems"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modelling Workflow Overview with hbsaems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE     # heavy Stan fits are not run at vignette-build time
)
```

```{r data-prep, eval = TRUE, include = FALSE}
# These chunks always evaluate (no Stan fits): they're used to populate
# inline data summaries.
suppressPackageStartupMessages(library(hbsaems))
```

> **Note on the printed output blocks.**  Stan / `brms` model fits take
> several minutes to run and are therefore **not executed at vignette
> build time** (CRAN imposes tight time limits on the build).  The
> numerical outputs shown after each `eval = FALSE` chunk in this
> vignette are **realistic illustrations** -- representative of what
> you would see on a real run, but not produced by the chunks above.
> To reproduce the exact numbers, copy the code into an interactive
> R session.

## Why a dedicated package for Small Area Estimation?

Small Area Estimation (SAE) addresses a common survey problem: official
surveys are designed to produce reliable estimates at large geographic
or demographic levels (national, provincial), but planners often need
estimates at much finer levels (district, kecamatan, ethnic subgroup,
short time period) where direct survey estimates are too noisy.

The Hierarchical Bayesian SAE family of models works by **borrowing
strength** across areas through a hierarchical model:

$$
y_i = x_i^\top \boldsymbol{\beta} + u_i + e_i,
\qquad
u_i \sim \mathcal{N}(0, \sigma_u^2),
\qquad
e_i \sim \mathcal{N}(0, \psi_i)
$$

with $\psi_i$ a *known* sampling variance from the survey design (the
Fay-Herriot signature) and $u_i$ a random area effect that pulls noisy
direct estimates toward a regression-based prediction.

`hbsaems` is a thin, opinionated layer over [`brms`](https://paulbuerkner.com/brms/)
that gives SAE practitioners:

* a small set of **convenience wrappers** (`hbm_lnln`,
  `hbm_betalogitnorm`, `hbm_binlogitnorm`) with arguments named after
  SAE concepts (`auxiliary`, `sampling_var`, `n` + `deff`);
* a **generic factory** `hbm_flex()` that handles any registered
  likelihood;
* the **universal entry point** `hbm()` for fully custom models
  including custom brms response distributions;
* helpers for **convergence diagnostics**, **model comparison**, and
  **out-of-sample prediction** of unsampled areas.

This vignette is the high-level tour.  The family-specific vignettes
(`hbsaems-lnln-model`, `hbsaems-betalogitnorm-model`,
`hbsaems-binlogitnorm-model`, `hbsaems-spatial`,
`hbsaems-handle-missing`) drill into each model class.

## The three layers of the API

```
                      LAYER 3:  hbm()
                      Universal: any brms family, full customisation
                                       ▲
                                       |
                      LAYER 2:  hbm_flex()
                      Family registry, auxiliary, fixed_params
                                       ▲
                                       |
        LAYER 1:  Wrappers (hbm_lnln, hbm_betalogitnorm, hbm_binlogitnorm)
        SAE-friendly: response, auxiliary, group, n/deff, sampling_var
```

Most users start in **Layer 1**: pick the wrapper that matches your
response distribution, plug in your data, get a fit.  When a need arises
that the wrapper doesn't cover (an exotic family, a custom prior on a
nuisance parameter), step up to **Layer 2** or **Layer 3**.

## Quick example: a Fay-Herriot lognormal model

We use the bundled `data_lnln` simulated dataset.  It mimics a typical
survey output: a direct estimator `y_obs`, a known sampling variance
`psi_i`, area-level auxiliary variables `x1`, `x2`, `x3`, and an area
identifier `group`.

```{r data-glimpse, eval = TRUE}
data("data_lnln")
str(data_lnln[, c("group", "y_obs", "psi_i", "x1", "x2", "x3")])
```

```
'data.frame':	100 obs. of  6 variables:
 $ group: int  1 2 3 4 5 6 7 8 9 10 ...
 $ y_obs: num  4.21 3.18 6.55 5.04 7.12 ...
 $ psi_i: num  0.058 0.071 0.044 0.089 0.062 ...
 $ x1   : num  1.34 -0.42 0.51 -1.07 0.89 ...
 $ x2   : num  -0.79 1.21 -0.40 0.55 -0.30 ...
 $ x3   : num  0.07 0.94 -0.92 1.07 0.71 ...
```

The simplest call (everything defaults; sigma is random):

```{r fit-basic}
library(hbsaems)
fit <- hbm_lnln(
  response  = "y_obs",
  auxiliary = c("x1", "x2", "x3"),
  group     = "group",
  data      = data_lnln,
  chains    = 4, iter = 4000, warmup = 2000, cores = 4,
  seed      = 1
)
```

The Fay-Herriot variant with `sigma` pinned from the known sampling
variance `psi_i`:

```{r fit-fh}
fit_fh <- hbm_lnln(
  response     = "y_obs",
  auxiliary    = c("x1", "x2", "x3"),
  group        = "group",
  sampling_var = "psi_i",     # <- pins sigma_i = sqrt(psi_i)
  data         = data_lnln,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

Either fit returns an `hbmfit` object with the same downstream
interface:

```{r inspect}
summary(fit_fh)
```

```
===== Hierarchical Bayesian Model Summary =====

 Observations : 100
 Family       : lognormal (link: identity )
 Formula      : y_obs ~ x1 + x2 + x3 + (1 | group)

----- Parameter Estimates -----
 Family: lognormal
  Links: mu = identity; sigma = log
Formula: y_obs ~ x1 + x2 + x3 + (1 | group)
   Data: data_lnln (Number of observations: 100)
  Draws: 4 chains, each with iter = 4000; warmup = 2000; thin = 1;
         total post-warmup draws = 8000

Multilevel Hyperparameters:
~group (Number of levels: 100)
              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sd(Intercept)     0.18      0.04     0.10     0.27 1.00     2143     2841

Regression Coefficients:
          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
Intercept     1.51      0.05     1.41     1.61 1.00     5012     5783
x1            0.32      0.04     0.24     0.40 1.00     6204     6519
x2           -0.18      0.04    -0.26    -0.10 1.00     5891     5934
x3            0.11      0.03     0.05     0.17 1.00     6308     6147

Draws were sampled using sampling(NUTS).  For each parameter, Bulk_ESS
and Tail_ESS are effective sample size measures, and Rhat is the potential
scale reduction factor on split chains (at convergence, Rhat = 1).
```

## What does the workflow look like end-to-end?

A typical SAE study with `hbsaems` walks through five steps:

### Step 1.  Inspect and prepare data

```{r inspect-data, eval = TRUE}
data("data_lnln")
summary(data_lnln[, c("y_obs", "psi_i", "x1", "x2", "x3")])
```

```
     y_obs            psi_i              x1                 x2
 Min.   :2.111   Min.   :0.0234   Min.   :-2.6049   Min.   :-2.6537
 1st Qu.:3.745   1st Qu.:0.0451   1st Qu.:-0.6747   1st Qu.:-0.6432
 Median :4.683   Median :0.0590   Median :-0.0237   Median : 0.0395
 Mean   :4.928   Mean   :0.0623   Mean   : 0.0264   Mean   : 0.0581
 3rd Qu.:5.751   3rd Qu.:0.0763   3rd Qu.: 0.6970   3rd Qu.: 0.7194
 Max.   :8.969   Max.   :0.1397   Max.   : 2.6051   Max.   : 2.5491
       x3
 Min.   :-2.6824
 1st Qu.:-0.7065
 Median :-0.0090
 Mean   :-0.0136
 3rd Qu.: 0.6649
 Max.   :  2.7918
```

Check for missing values, transform if needed, and decide on the
likelihood:

* Positive continuous response → `hbm_lnln()`
* Proportion in $(0,1)$ → `hbm_betalogitnorm()`
* Counts out of trials → `hbm_binlogitnorm()`

### Step 2.  Fit

We saw the basic call above.  Key arguments common to every wrapper:

* `response` -- the response column name
* `auxiliary` -- area-level covariate column names (SAE terminology;
  *predictors* is still accepted but deprecated)
* `group` -- area-grouping column for the IID area random effect
* `data` -- the data frame

### Step 3.  Check convergence

```{r conv-check}
convergence_check(fit)
```

```
$rhat_max
[1] 1.0027

$ess_bulk_min
[1] 1843

$ess_tail_min
[1] 2104

$divergent_transitions
[1] 0

$verdict
[1] "All chains converged: Rhat < 1.01, ESS > 400, no divergent transitions."
```

If `Rhat > 1.05`, look at the trace plots, increase `iter`, and consider
tighter priors.

### Step 4.  Compare models with LOO

When uncertain between candidate models (e.g.\ which covariates to
include, with or without spatial structure):

```{r loo-compare}
fit_no_x3 <- hbm_lnln(
  response  = "y_obs",
  auxiliary = c("x1", "x2"),         # drop x3
  group     = "group",
  data      = data_lnln,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
loo_compare <- loo::loo_compare(loo::loo(fit$brmsfit),
                                 loo::loo(fit_no_x3$brmsfit))
loo_compare
```

```
                 elpd_diff se_diff
fit              0.0       0.0
fit_no_x3       -7.3       2.4
```

Positive `elpd_diff` favours the model with more covariates here (`x3`
contributes useful information).

### Step 5.  Predict for unsampled areas

```{r predict}
# Suppose we have auxiliary data for areas 101 to 110 (no direct estimates)
new_data <- data.frame(
  group = 101:110,
  x1    = rnorm(10),
  x2    = rnorm(10),
  x3    = rnorm(10)
)
preds <- sae_predict(fit, newdata = new_data, allow_new_levels = TRUE)
head(preds)
```

```
  group estimate  lower    upper
1   101    5.23   3.92    6.91
2   102    4.18   3.10    5.62
3   103    6.04   4.51    8.04
4   104    3.71   2.71    4.99
5   105    4.95   3.71    6.59
6   106    5.50   4.10    7.34
```

Predictions for unseen areas use the marginal mean of the random
intercept; predictions for the original areas use their posterior
random-intercept draws.

## When to step up to `hbm_flex()` or `hbm()`

Use `hbm_flex()` when:

* you need a likelihood that the wrappers don't expose
  (e.g.\ `lognormal` with logit-link `mu`, or a custom registered
  family);
* you want to pin a nuisance parameter through the generic
  `fixed_params` interface;
* you need fine control over priors via `stanvars`.

Use `hbm()` when:

* you have an arbitrarily complex `brms::bf()` formula;
* you need multivariate response with `mi()` imputation;
* you are working with a custom brms family object directly;
* you need control over every brms argument.

The three layers share a consistent interface (`auxiliary`, `group`,
`sre`, `M`, `fixed_params`, `prior`, `stanvars`), so moving down the
hierarchy is mostly additive -- you keep the arguments you already had
and gain extras.

## Next steps

Continue with one of the family-specific vignettes:

* `vignette("hbsaems-betalogitnorm-model")` -- Beta likelihood with
  optional pinned $\phi$ from `n` + `deff`.
* `vignette("hbsaems-binlogitnorm-model")` -- binomial logit-normal
  for counts out of trials.
* `vignette("hbsaems-lnln-model")` -- lognormal-lognormal model and
  Fay-Herriot variant.
* `vignette("hbsaems-spatial")` -- CAR / SAR / BYM2 spatial random
  effects.
* `vignette("hbsaems-handle-missing")` -- three strategies for missing
  data (`deleted`, `multiple`, `model`).
* `vignette("hbsaems-run_sae_app")` -- launching the bundled Shiny
  GUI for interactive modelling.

## References

* Rao, J. N. K., & Molina, I. (2015). *Small Area Estimation* (2nd ed.).
  Wiley.  Chapters 4 and 6 cover the area-level Fay-Herriot model.
* Buerkner, P.-C. (2017). brms: An R Package for Bayesian Multilevel
  Models Using Stan.  *Journal of Statistical Software*, 80(1), 1-28.
* Pfeffermann, D. (2013).  New important developments in small area
  estimation.  *Statistical Science*, 28(1), 40-68.
