---
title: "Beta Logit-Normal SAE Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beta Logit-Normal SAE Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

```{r preload, eval = TRUE, include = FALSE}
suppressPackageStartupMessages(library(hbsaems))
```

> **Note on the printed output blocks.**  Stan / `brms` model fits take
> several minutes to run and are therefore **not executed at vignette
> build time**.  The numerical outputs shown after each `eval = FALSE`
> chunk in this vignette are **realistic illustrations** -- representative
> of what you would see on a real run, but not produced by the chunks
> above.  To reproduce the exact numbers, copy the code into an
> interactive R session.

## When to use the Beta likelihood for SAE

Many SAE problems involve a response that is a **proportion**, strictly
between 0 and 1: literacy rate, vaccine coverage, poverty headcount
ratio, employment-to-population ratio.  A Gaussian linear model on the
raw response is conceptually wrong (it can predict values outside
$[0,1]$) and a Gaussian model on the logit-transformed response can be
poor when the true proportion is near 0 or 1.

The Beta logit-normal model is the standard alternative
(Liu 2009; Rao & Molina 2015, p.\ 390):

$$
\begin{aligned}
y_i \mid \theta_i, \phi_i
   &\sim \mathrm{Beta}\!\bigl(\theta_i \phi_i, \, (1 - \theta_i)\phi_i\bigr), \\
\mathrm{logit}(\theta_i)
   &= x_i^\top \boldsymbol{\beta} + u_i, \quad
     u_i \sim \mathcal{N}(0, \sigma_u^2),
\end{aligned}
$$

where $\theta_i \in (0,1)$ is the area mean (parameterised on the logit
scale) and $\phi_i > 0$ is a precision parameter controlling the
within-area dispersion.  A larger $\phi_i$ means less spread.

For survey-design data, $\phi_i$ has an information-theoretic
interpretation:

$$
\phi_i = \frac{n_i}{\mathrm{deff}_i} - 1,
$$

so the precision is large when the effective sample size $n_i /
\mathrm{deff}_i$ is large.  `hbsaems` lets you either **pin** $\phi_i$
to this value or **sample** it with a hyperprior.

## Three modes for $\phi$

`hbm_betalogitnorm()` recognises three distinct workflows:

| Mode | Trigger | Description |
|------|---------|-------------|
| Random | (default) | $\phi \sim \mathrm{Gamma}(\alpha, \beta)$, with $\alpha, \beta$ themselves given priors |
| Fixed | `n = "n"` + `deff = "deff"` | $\phi_i = n_i / \mathrm{deff}_i - 1$ pinned via offset |
| Custom hyperprior | `stanvars = ...` | User overrides the default $\mathrm{Gamma}(1,1)$ priors on $\alpha, \beta$ |

A fourth mode using the **generic `fixed_params` interface** is also
available for power users (see the last section).

## The data

The bundled `data_betalogitnorm` has a typical SAE-from-survey
structure:

```{r show-data, eval = TRUE}
data("data_betalogitnorm")
str(data_betalogitnorm[, c("regency", "y", "n", "deff", "x1", "x2", "x3")])
```

```
'data.frame':	50 obs. of  7 variables:
 $ regency: chr  "regency_001" "regency_002" "regency_003" "regency_004" ...
 $ y    : num  0.452 0.371 0.624 0.518 0.401 ...
 $ n    : int  120 95 142 108 87 153 124 102 91 138 ...
 $ deff : num  1.31 1.08 1.42 1.19 1.05 ...
 $ x1   : num  0.82 -0.41 1.04 0.15 -0.92 ...
 $ x2   : num  -0.31 0.74 -0.55 0.21 1.04 ...
 $ x3   : num  0.18 0.42 0.07 -0.31 0.65 ...
```

The response `y` is the direct survey estimate of an area-level
proportion; `n` and `deff` come from the survey microdata; `x1`, `x2`,
`x3` are area-level auxiliaries from administrative or census sources.

## Mode 1: random $\phi$ (default)

The simplest call lets $\phi$ be sampled with a hyperprior:

```{r fit-random}
library(hbsaems)
library(brms)

fit_random <- hbm_betalogitnorm(
  response  = "y",
  auxiliary = c("x1", "x2", "x3"),
  area_var   = "regency",
  data      = data_betalogitnorm,
  chains = 4, iter = 4000, warmup = 2000, cores = 4,
  seed = 1
)
summary(fit_random)
```

```
===== Hierarchical Bayesian Model Summary =====

 Observations : 50
 Family       : Beta (link: logit )
 Formula      : y ~ x1 + x2 + x3 + (1 | regency)

----- Parameter Estimates -----
 Family: beta
  Links: mu = logit; phi = identity
Formula: y ~ x1 + x2 + x3 + (1 | regency)
   Data: data_betalogitnorm (Number of observations: 50)
  Draws: 4 chains, each with iter = 4000; warmup = 2000; thin = 1;
         total post-warmup draws = 8000

Multilevel Hyperparameters:
~group (Number of levels: 50)
              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sd(Intercept)     0.21      0.06     0.10     0.33 1.00     1842     2615

Regression Coefficients:
          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
Intercept    -0.18      0.07    -0.31    -0.05 1.00     4527     5394
x1            0.36      0.04     0.28     0.43 1.00     6308     6541
x2           -0.21      0.04    -0.29    -0.13 1.00     5912     6105
x3            0.08      0.04     0.00     0.16 1.00     6184     6027

Further Distributional Parameters:
       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
phi      62.84     12.71    41.22    91.40 1.00     2841     3217
alpha     1.45      0.92     0.10     3.64 1.00     2104     2519
beta      0.04      0.03     0.00     0.11 1.00     1873     2241

Draws were sampled using sampling(NUTS).
```

Note that `alpha` and `beta` (the hyperparameters of the Gamma prior on
$\phi$) appear in the output because they are sampled in this mode.
The default hyperpriors are
$\alpha \sim \mathrm{Gamma}(1,1)$ and $\beta \sim \mathrm{Gamma}(1,1)$.

> **Hyperprior choice for SAE applications.**  The default
> $\mathrm{Gamma}(1,1)$ priors place most prior mass on $\phi \approx
> 0.5\text{--}5$, which corresponds to very high within-area
> variability.  Real survey proportions typically have $\phi$ in the
> range 10--100.  If your fit produces a posterior mean of $\phi$
> uncomfortably close to the default prior mean, consider widening the
> hyperpriors via the `stanvars` argument (see *Mode 3* below) -- for
> example, $\alpha \sim \mathrm{Gamma}(0.1, 0.1)$ and
> $\beta \sim \mathrm{Gamma}(0.1, 0.1)$ for a weakly-informative
> hierarchy.  Alternatively, if you have access to survey design
> information, use **Mode 2** (pinned $\phi$) -- generally preferred
> in SAE practice.

## Mode 2: $\phi$ pinned from survey design

When the survey microdata are available, pinning $\phi_i$ to the
known precision $n_i / \mathrm{deff}_i - 1$ is more informative and
typically gives **smaller posterior intervals** for $\theta_i$:

```{r fit-fixed}
fit_fixed <- hbm_betalogitnorm(
  response  = "y",
  auxiliary = c("x1", "x2", "x3"),
  n         = "n",                   # <- column with sample sizes
  deff      = "deff",                # <- column with design effects
  area_var   = "regency",
  data      = data_betalogitnorm,
  chains = 4, iter = 4000, warmup = 2000, cores = 4,
  seed = 1
)
summary(fit_fixed)
```

```
===== Hierarchical Bayesian Model Summary =====

 Observations : 50
 Family       : Beta (link: logit )
 Formula      : y ~ x1 + x2 + x3 + (1 | regency)
                phi ~ 0 + offset(.hbsaems_phi_fixed)

----- Parameter Estimates -----
 Family: beta
  Links: mu = logit; phi = identity
Formula: y ~ x1 + x2 + x3 + (1 | regency)
   Data: data_betalogitnorm (Number of observations: 50)
  Draws: 4 chains, each with iter = 4000; warmup = 2000; thin = 1;
         total post-warmup draws = 8000

Multilevel Hyperparameters:
~group (Number of levels: 50)
              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sd(Intercept)     0.18      0.04     0.10     0.27 1.00     2843     3415

Regression Coefficients:
          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
Intercept    -0.16      0.05    -0.25    -0.07 1.00     5841     6024
x1            0.34      0.03     0.28     0.40 1.00     7158     6841
x2           -0.20      0.03    -0.26    -0.14 1.00     6915     6517
x3            0.09      0.03     0.03     0.15 1.00     7204     6624

Draws were sampled using sampling(NUTS).
```

Observations:

* The `phi`, `alpha`, `beta` lines from Mode 1 are **gone** -- those
  parameters are no longer sampled.
* The CIs on the regression coefficients are **tighter** (e.g.\ x1:
  $[0.28, 0.40]$ vs $[0.28, 0.43]$ in Mode 1), since $\phi$ uncertainty
  is no longer propagated.

## Mode 3: custom hyperprior on $\alpha, \beta$

If the default $\mathrm{Gamma}(1,1)$ priors on $\alpha$ and $\beta$ are
too vague (or too tight) for your data, override them via `stanvars`:

```{r fit-custom-hyperprior}
fit_custom <- hbm_betalogitnorm(
  response  = "y",
  auxiliary = c("x1", "x2", "x3"),
  area_var   = "regency",
  data      = data_betalogitnorm,
  stanvars  = stanvar(scode = "alpha ~ gamma(2, 1);", block = "model") +
              stanvar(scode = "beta  ~ gamma(2, 3);", block = "model"),
  chains = 4, iter = 4000, warmup = 2000, cores = 4,
  seed = 1
)
summary(fit_custom)
```

```
Further Distributional Parameters:
       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
phi      54.91      9.83    37.65    76.20 1.00     3142     3674
alpha     1.92      0.81     0.59     3.78 1.00     2871     3093
beta      0.08      0.04     0.02     0.18 1.00     2517     2841
```

The user-supplied $\mathrm{Gamma}(2, 1)$ on $\alpha$ shifts its posterior
toward higher values, which in turn changes the posterior of $\phi$.
You can supply only one of the two and the other will retain its
default; you can also pin one to a specific value via
`stanvar(scode = "alpha = 1.5;", block = "tparameters")`.

You can also override the prior on $\phi$ itself (not the hyperpriors)
via the `prior` argument:

```{r fit-phi-prior}
fit_phi <- hbm_betalogitnorm(
  response  = "y",
  auxiliary = c("x1", "x2", "x3"),
  area_var   = "regency",
  data      = data_betalogitnorm,
  prior     = brms::set_prior("gamma(2, 0.05)", class = "phi"),
  chains = 4, iter = 4000, warmup = 2000, cores = 4,
  seed = 1
)
```

When `prior` overrides `class = "phi"`, the `stanvars` machinery is
not used and the default `Intercept ~ student_t(4, 0, 10)` and
`b ~ student_t(4, 0, 2.5)` are still filled in for the other classes.

> **Conflict policy.**  When $\phi$ is pinned (Mode 2 or Mode 4 below),
> it is no longer a Stan parameter -- it is fixed in the model
> formula via an offset.  Supplying a `prior` on `phi` or a
> `stanvars` sampling statement targeting `phi`, `alpha`, or `beta`
> in this situation is a logical contradiction and is rejected with
> a clear error at construction time.  This prevents silent surprises
> where the user thinks they have specified a prior that is actually
> ignored.

## Mode 4: generic `fixed_params` (power user)

The wrapper-specific `n` and `deff` arguments are syntactic sugar over
a generic `fixed_params` mechanism that works for **any** parameter
of **any** family.  Equivalent of Mode 2:

```{r fit-flex}
fit_flex <- hbm_flex(
  family_key   = "Beta",
  response     = "y",
  auxiliary    = c("x1", "x2", "x3"),
  area_var   = "regency",
  data         = data_betalogitnorm,
  fixed_params = list(phi = ~ I(n / deff - 1)),    # <- formula spec
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

The `fixed_params` argument accepts four kinds of specifications -- see
`?hbm` for the full table.  This pattern also works for custom
distributions registered via `register_hbsae_brms_custom()`.

## Checking convergence

For every fit, the standard diagnostic check is:

```{r conv-check}
convergence_check(fit_fixed)
```

```
$rhat_max
[1] 1.0019

$ess_bulk_min
[1] 2843

$ess_tail_min
[1] 3415

$divergent_transitions
[1] 0

$verdict
[1] "All chains converged: Rhat < 1.01, ESS > 400, no divergent transitions."
```

If divergences appear, raise `adapt_delta` via
`control = list(adapt_delta = 0.99)` and consider tighter priors on
$\sigma_u$.

## Comparing modes with LOO

When the survey microdata are available, it is good practice to
compare Modes 1 and 2 to see whether pinning $\phi$ is supported by
the data:

```{r loo-compare}
loo_compare <- loo::loo_compare(
  loo::loo(fit_random$brmsfit),
  loo::loo(fit_fixed$brmsfit)
)
loo_compare
```

```
                  elpd_diff se_diff
fit_fixed         0.0       0.0
fit_random       -2.1       1.7
```

A small `elpd_diff` (less than 2 SE) suggests both modes fit the data
comparably; pick the simpler (fixed-$\phi$) model for inference.

## Prediction for new areas

```{r predict}
new_areas <- data.frame(
  group = 51:55,
  x1    = rnorm(5),
  x2    = rnorm(5),
  x3    = rnorm(5)
)
preds <- sae_predict(fit_fixed, newdata = new_areas, allow_new_levels = TRUE)
preds
```

```
  group estimate  lower  upper
1    51    0.421  0.298  0.561
2    52    0.508  0.379  0.643
3    53    0.347  0.231  0.482
4    54    0.612  0.481  0.731
5    55    0.385  0.262  0.524
```

Predictions are reported on the *response* scale (proportions in
$(0,1)$), with 95% credible intervals.

## Summary: which mode should I use?

* **Mode 1 (random $\phi$)** -- use when `n` and `deff` are unknown or
  unreliable; produces wider posterior intervals.
* **Mode 2 (fixed $\phi$)** -- use when you have clean survey design
  information; produces tighter intervals.  *Recommended for standard
  SAE work*.
* **Mode 3 (custom hyperprior)** -- use when you have strong prior
  beliefs about the typical magnitude of $\phi$ across areas.
* **Mode 4 (generic `fixed_params`)** -- use through `hbm_flex()` when
  building unusual models or working with custom distributions.

## References

* Liu, B. (2009). *Hierarchical Bayes Estimation and Empirical Best
  Prediction of Small-Area Proportions*.  PhD thesis, University of
  Maryland.
* Rao, J. N. K., & Molina, I. (2015). *Small Area Estimation*, 2nd ed.
  Wiley.  See Chapter 8 for binary/Beta SAE.
