---
title: "Handling Missing Data in hbsaems"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Handling Missing Data in hbsaems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

```{r preload, eval = TRUE, include = FALSE}
suppressPackageStartupMessages(library(hbsaems))
```

> **Note on the printed output blocks.**  Stan / `brms` model fits take
> several minutes to run and are therefore **not executed at vignette
> build time**.  The numerical outputs shown after each `eval = FALSE`
> chunk in this vignette are **realistic illustrations** -- representative
> of what you would see on a real run, but not produced by the chunks
> above.  To reproduce the exact numbers, copy the code into an
> interactive R session.

## The Bayesian principle: response variables are never imputed

A core design rule in `hbsaems` is that **multiple imputation only ever
imputes auxiliary (predictor) variables, never the response**.  This is
not a limitation -- it is the statistically correct Bayesian treatment:

* For a Bayesian hierarchical model with response $Y$, the posterior
  for the parameters $\theta$ marginalises the missing outcomes
  $Y_{\text{mis}}$:
  $$
  p(\theta \mid Y_{\text{obs}}, X) = \int p(\theta \mid Y_{\text{obs}},
    Y_{\text{mis}}, X) \, p(Y_{\text{mis}} \mid Y_{\text{obs}}, X, \theta)
    \, \mathrm{d}Y_{\text{mis}}.
  $$
* Plugging in a mice-imputed value for $Y$ would replace this
  integral with a single point substitute, deflate the posterior
  uncertainty, and bias estimates if the imputation model is
  misspecified.
* The correct way to "fill in" missing $Y$ values is **joint
  Bayesian imputation** via `brms::mi()`, which samples $Y_{\text{mis}}$
  alongside $\theta$.  This is the `handle_missing = "model"`
  strategy described below.

The three strategies thus have **different scopes**:

| Strategy   | What it imputes                              | Required completeness |
|------------|----------------------------------------------|------------------------|
| `"deleted"`| Nothing -- drops rows with missing $Y$   | Auxiliary $X$ must be complete |
| `"multiple"`| **Only auxiliary $X$** via `mice`        | Any $Y$ missingness handled separately (see below) |
| `"model"`  | **$X$ and $Y$ jointly via `brms::mi()` | Continuous outcomes only |

When `handle_missing = NULL` (default), the package **auto-selects** a
strategy based on the family's `supports_mi` flag.

## Missing-data patterns in SAE

Three patterns are common:

1. **Missing response, complete covariates** -- some areas were not
   sampled at all, but you have administrative auxiliaries for them.
2. **Complete response, missing covariates** -- the survey reached
   every area, but a census variable is missing for some.
3. **Missing in both** -- a mix of the above.

The package's reaction to each pattern depends on which `handle_missing`
strategy is in effect.  The next three sections walk through each
strategy and the patterns it handles.

## Strategy 1: `"deleted"` (listwise deletion on Y only)

When `handle_missing = "deleted"`:

* Rows where the **response $Y$ is missing** are dropped before
  model fitting.
* If any **auxiliary** is missing, the function **refuses with an
  informative error** -- listwise deletion of $X$ would discard
  information that more principled strategies can recover.

```{r delete}
library(hbsaems)
data("data_fhnorm")

# Pattern 1: missing Y only, X complete
data_miss_y      <- data_fhnorm
data_miss_y$y[c(3, 14, 27)] <- NA

fit_deleted <- hbm(
  formula        = brms::bf(y ~ x1 + x2 + x3),
  re             = ~ (1 | regency),
  data           = data_miss_y,
  handle_missing = "deleted",
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

```
handle_missing = 'deleted': 3 row(s) with missing response variable removed
from model fitting.
```

If you pass data with missing $X$ to `"deleted"`, the call is
rejected:

```{r delete-error}
data_miss_x <- data_fhnorm
data_miss_x$x1[6:8] <- NA

hbm(brms::bf(y ~ x1 + x2 + x3),
    data           = data_miss_x,
    re             = ~ (1 | regency),
    handle_missing = "deleted")
```

```
Error: Option `handle_missing = 'deleted'` requires all auxiliary (predictor)
variables to be complete. Missing values were detected in: x1. Consider
`handle_missing = 'multiple'` to impute missing predictors.
```

The dropped rows (under the legitimate Y-only case) are retained in
`fit_deleted$data` so that `sae_predict()` can still produce
predictions for those areas using their auxiliary values.

## Strategy 2: `"multiple"` (mice on X only)

When `handle_missing = "multiple"`, `mice` imputes the missing
**auxiliary variables** $m$ times, fits the brms model on each
imputed dataset, and pools the posterior draws via Rubin's rules (via
`brms::brm_multiple`).

The response is **never imputed by mice**.  How the package handles
each pattern:

### Pattern 2a: Missing X only -- normal mice workflow

```{r mi-x}
data_miss_x         <- data_fhnorm
data_miss_x$x1[6:8] <- NA

fit_mi <- hbm(
  formula        = brms::bf(y ~ x1 + x2 + x3),
  re             = ~ (1 | regency),
  data           = data_miss_x,
  handle_missing = "multiple",
  m              = 5,                       # 5 imputations
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

```
Missing predictor variable(s): x1. Applying multiple imputation (mice)
with m = 5 imputations.

Compiling the C++ model
Fitting imputed model 1
Start sampling
...
Fitting imputed model 5
Start sampling
...

(Posterior draws pooled via Rubin's rules.)
```

The pooled posterior is returned in the usual `hbmfit` interface.
Diagnostics are *per-imputation*; the global Rhat / ESS view is not
informative for pooled chains -- see `?brm_multiple` for details.

### Pattern 2b: Missing X *and* Y (or only Y) -- response not imputed

If $Y$ also has missing values, the package issues a warning
explaining that **mice will not impute $Y$** and the rows with
missing $Y$ are excluded from fitting but retained in the returned
object:

```{r mi-y-warning}
data_miss_both <- data_fhnorm
data_miss_both$y[2:3]   <- NA           # missing Y
data_miss_both$x1[6:8]  <- NA           # missing X

fit_mi_both <- hbm(
  formula        = brms::bf(y ~ x1 + x2 + x3),
  re             = ~ (1 | regency),
  data           = data_miss_both,
  handle_missing = "multiple"
)
```

```
Warning: Missing values detected in response variable(s): y.
Multiple imputation via `mice` applies ONLY to auxiliary predictor
variables (X). Response variable(s) are NEVER imputed inside a
Bayesian model -- missing Y rows will be excluded from model fitting
but are retained in the returned object for prediction via hbsae().
If you wish to model missingness in Y jointly with the parameters,
consider `handle_missing = 'model'` (continuous outcomes only).

Missing predictor variable(s): x1. Applying multiple imputation (mice)
with m = 5 imputations.
```

### Pattern 2c: Missing Y only -- auto-conversion

If `handle_missing = "multiple"` is given but **only Y** is missing
($X$ is complete), there is nothing for mice to impute.  The
package then auto-converts to a more appropriate strategy:

* **Continuous family** (e.g.\ `gaussian`, `lognormal`):
  auto-converts to `handle_missing = "model"` and adds `| mi()` to the
  formula LHS so that missing $Y$ values are jointly sampled with
  the parameters (the correct Bayesian approach).
* **Discrete family** (e.g.\ `binomial`, `poisson`):
  `mi()` is not supported -- falls back to `handle_missing = "deleted"`,
  dropping the rows with missing $Y$.

```{r mi-y-only-continuous}
data_miss_y_only <- data_fhnorm
data_miss_y_only$y[2:3] <- NA           # only Y missing

# Gaussian family -> auto-converts to "model" with | mi() on LHS
fit_auto1 <- hbm(
  formula        = brms::bf(y ~ x1 + x2 + x3),
  re             = ~ (1 | regency),
  data           = data_miss_y_only,
  handle_missing = "multiple"
)
```

```
handle_missing = 'multiple' was specified but only the response variable
(Y) is missing and X is complete. mice imputes predictor variables only and
cannot impute Y. Automatically converting to handle_missing = 'model':
'| mi()' will be added to the formula LHS so that brms jointly estimates
missing Y with the model parameters (the correct Bayesian approach).
```

### Controlling mice

Pass options through `mice_args`:

```{r mi-args}
fit_mi2 <- hbm(
  formula        = brms::bf(y ~ x1 + x2 + x3),
  re             = ~ (1 | regency),
  data           = data_miss_x,
  handle_missing = "multiple",
  m              = 10,
  mice_args      = list(method = "pmm", maxit = 20),
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

## Strategy 3: `"model"` (joint Bayesian imputation)

When `handle_missing = "model"`, missing values in both response and
auxiliary variables are sampled jointly with the model parameters via
`brms::mi()`.  The user must specify imputation sub-models explicitly
in the formula:

```{r mi-model}
data_miss_both <- data_fhnorm
data_miss_both$y[2:3]   <- NA
data_miss_both$x1[6:8]  <- NA

fit_model <- hbm(
  formula        = brms::bf(y  | mi() ~ mi(x1) + x2 + x3) +
                   brms::bf(x1 | mi() ~ x2 + x3),
  re             = ~ (1 | regency),
  data           = data_miss_both,
  handle_missing = "model",
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

```
Compiling Stan program...
Start sampling...
```

* `y | mi() ~ ...` tells brms to treat missing $Y$ values as
  parameters to be sampled.
* `mi(x1)` on the RHS tells brms to use the imputed $x_1$ (also
  sampled) as a predictor.
* The second `bf(x1 | mi() ~ x2 + x3)` is the imputation sub-model for
  $x_1$.

The result is a **multivariate brms model** in which all missing
values are sampled alongside the regression coefficients; posterior
uncertainty in the imputations is automatically propagated.

**Restriction**: `handle_missing = "model"` only works for **continuous
distributions** (the `supports_mi` flag in the family registry).  For
discrete families (binomial, Poisson, ...), `brms::mi()` is not
supported -- use `"multiple"` for missing $X$ and accept listwise
deletion for missing $Y$.

## Auto-selection (when `handle_missing = NULL`)

If you omit `handle_missing`, the package looks at whether **any**
missingness exists and selects based on the family's `supports_mi`
flag:

```{r auto}
# Build a data frame with some missing values to demonstrate
data("data_lnln")
data_with_some_na <- data_lnln
data_with_some_na$x1[c(3, 14, 27)] <- NA    # missing covariate

fit_auto <- hbm_lnln(
  response  = "y_obs",
  auxiliary = c("x1", "x2", "x3"),
  area_var  = "district",
  data      = data_with_some_na,
  # handle_missing not given
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

```
Missing values detected. Auto-selecting handle_missing = 'multiple'
(family 'lognormal', supports_mi = FALSE).
```

The actual decision rule used in the code is:

```
Any NA detected in response or auxiliary?
  └── family.supports_mi == TRUE?   ->  "model"   (joint mi(); user must
  │                                                supply mi() formula
  │                                                terms when missing X)
  └── otherwise                      ->  "multiple" (mice on X only;
                                                     missing Y handled
                                                     per Strategy 2 logic)
```

The auto-selection is intentionally simple and does **not** branch on
"Y-only" vs "X-only" patterns: those subtleties are handled inside the
chosen strategy (see Pattern 2c above for the case where `"multiple"`
detects Y-only missingness and auto-converts to `"model"` for
continuous families or to `"deleted"` for discrete families).

You can always override by passing an explicit
`handle_missing = "..."`.

## Comparing strategies (when fitting all three is feasible)

```{r compare}
loo_compare <- loo::loo_compare(
  loo::loo(fit_deleted$brmsfit),
  loo::loo(fit_mi$brmsfit),
  loo::loo(fit_model$brmsfit)
)
loo_compare
```

```
                  elpd_diff se_diff
fit_model         0.0       0.0
fit_mi           -2.1       2.4
fit_deleted     -10.7       3.8
```

Note that `fit_deleted` and `fit_mi` here refer to fits on the same
pattern (missing Y only, missing X only respectively), so the
comparison is over comparable likelihood scales.

## Practical guidance

* **Small fraction missing (< 5%) and likely MCAR, missing in Y only**
  -- `"deleted"` is fine.
* **Missing in X (with or without missing Y)**:
  - Continuous family -- prefer `"model"` (joint Bayesian imputation,
    most principled).
  - Discrete family -- use `"multiple"` (mice on X; rows with missing
    Y are dropped from fitting but retained in the result for
    prediction).
* **Missing in response only, you have administrative auxiliaries for
  those areas** -- prefer `"deleted"` plus `sae_predict()` (this is
  more transparent than imputing the response).
* **Reproducibility** -- set `seed` for both the wrapper call and (if
  using mice) within `mice_args` to make imputations reproducible.

## References

* Rubin, D. B. (1987).  *Multiple Imputation for Nonresponse in
  Surveys*.  Wiley.
* van Buuren, S. (2018).  *Flexible Imputation of Missing Data*.
  CRC Press.
* Bürkner, P.-C. (2017).  brms: An R Package for Bayesian Multilevel
  Models Using Stan.  *Journal of Statistical Software*, 80(1),
  1-28.  Section on `mi()`.
* Gelman, A., & Hill, J. (2006).  *Data Analysis Using Regression
  and Multilevel/Hierarchical Models*.  Cambridge University Press,
  Chapter 25.
