---
title: "hbsaems: Complete Examples Reference"
author: "Achmad Syahrul Choir, Saniyyah Sri Nurhayati, Sofi Zamzanah, Arsyka Laila Oktalia Siregar"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: paged
    highlight: tango
    theme: cosmo
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE,        # All Stan fits not evaluated at build time
  message  = FALSE,
  warning  = FALSE
)
```

```{r preload, eval = TRUE, include = FALSE}
suppressPackageStartupMessages({
  library(hbsaems)
})
```

> **Note on the printed output blocks.**  Stan / `brms` model fits take
> several minutes to run and are therefore **not executed at document
> build time** -- CRAN imposes tight time limits on R package documents.
> The numerical outputs shown after each `eval = FALSE` chunk are
> **realistic illustrations**: representative of what you would see on
> a real run, but not produced by the chunks above.  To reproduce the
> exact numbers, copy the code into an interactive R session.

---

# Introduction

This document is a **comprehensive examples reference** for the
**hbsaems** package, covering every supported model family, every
advanced feature, and every utility in the Bayesian workflow.

The package fits **area-level** Hierarchical Bayesian Small Area
Estimation (HBSAE) models following the methodological tradition of
Rao and Molina (2015), adapted to the parameterisation and
prior-specification conventions of the [brms](https://paulbuerkner.com/brms/)
package (Bürkner, 2017), which targets the Stan back-end.

For a topical tour, see the dedicated vignettes:

* `vignette("hbsaems-modelling")` -- high-level workflow overview
* `vignette("hbsaems-lnln-model")` -- lognormal-lognormal model
* `vignette("hbsaems-betalogitnorm-model")` -- beta logit-normal
* `vignette("hbsaems-binlogitnorm-model")` -- binomial logit-normal
* `vignette("hbsaems-spatial")` -- CAR / SAR / BYM2
* `vignette("hbsaems-handle-missing")` -- missing data strategies
* `vignette("hbsaems-run_sae_app")` -- Shiny application

---

# 1. Quick-Start Examples by Family

## 1.1 Fay-Herriot Normal model

The classic area-level model with a Gaussian likelihood:

$$
\begin{aligned}
y_i \mid \theta_i, \psi_i
   &\sim \mathcal{N}(\theta_i, \psi_i), \quad \psi_i \text{ known} \\
\theta_i
   &= x_i^\top \boldsymbol{\beta} + u_i, \quad
     u_i \sim \mathcal{N}(0, \sigma_u^2)
\end{aligned}
$$

```{r fh-normal}
library(hbsaems)
library(brms)

data("data_fhnorm")
head(data_fhnorm)

fit_fh <- hbm(
  formula     = bf(y ~ x1 + x2 + x3),
  hb_sampling = "gaussian",
  hb_link     = "identity",
  re          = ~(1 | group),
  data        = data_fhnorm,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)

summary(fit_fh)
```

```
===== Hierarchical Bayesian Model Summary =====

 Observations : 100
 Family       : gaussian (link: identity )
 Formula      : y ~ x1 + x2 + x3 + (1 | group)

----- Parameter Estimates -----
Multilevel Hyperparameters:
~group (Number of levels: 100)
              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sd(Intercept)     0.31      0.05     0.21     0.42 1.00     2034     2841

Regression Coefficients:
          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
Intercept     1.72      0.09     1.55     1.89 1.00     4218     5104
x1            0.45      0.04     0.37     0.53 1.00     6418     6628
x2           -0.33      0.03    -0.39    -0.27 1.00     6021     6105
x3            0.19      0.02     0.15     0.23 1.00     6304     6207

Further Distributional Parameters:
      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sigma     0.21      0.02     0.17     0.25 1.00     5417     6028
```

## 1.2 Beta logit-normal: Four operational modes

Beta likelihood for proportions in $(0,1)$:

$$
\begin{aligned}
y_i \mid \theta_i, \phi_i
   &\sim \mathrm{Beta}(\theta_i \phi_i, (1-\theta_i)\phi_i) \\
\mathrm{logit}(\theta_i)
   &= x_i^\top \boldsymbol{\beta} + u_i
\end{aligned}
$$

```{r beta-data, eval = TRUE}
data("data_betalogitnorm")
str(data_betalogitnorm[, c("y", "n", "deff", "x1", "x2", "x3", "group")])
```

### Mode 1: Random precision $\phi$ (default)

```{r beta-mode1}
fit_beta_random <- hbm_betalogitnorm(
  response  = "y",
  auxiliary = c("x1", "x2", "x3"),
  group     = "group",
  data      = data_betalogitnorm,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
summary(fit_beta_random)
```

```
Further Distributional Parameters:
       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
phi      62.84     12.71    41.22    91.40 1.00     2841     3217
alpha     1.45      0.92     0.10     3.64 1.00     2104     2519
beta      0.04      0.03     0.00     0.11 1.00     1873     2241
```

### Mode 2: Pinned $\phi$ from survey design

$\phi_i = n_i / \mathrm{deff}_i - 1$ pinned via offset:

```{r beta-mode2}
fit_beta_fixed <- hbm_betalogitnorm(
  response  = "y",
  auxiliary = c("x1", "x2", "x3"),
  n         = "n",
  deff      = "deff",
  group     = "group",
  data      = data_betalogitnorm,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
summary(fit_beta_fixed)
```

```
Formula: y ~ x1 + x2 + x3 + (1 | group)
         phi ~ 0 + offset(.hbsaems_phi_fixed)

Regression Coefficients:
          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
Intercept    -0.16      0.05    -0.25    -0.07 1.00     5841     6024
x1            0.34      0.03     0.28     0.40 1.00     7158     6841
x2           -0.20      0.03    -0.26    -0.14 1.00     6915     6517
x3            0.09      0.03     0.03     0.15 1.00     7204     6624
```

### Mode 3: Custom hyperprior on $\alpha, \beta$

Override the default $\mathrm{Gamma}(1,1)$ priors:

```{r beta-mode3}
fit_beta_custom <- hbm_betalogitnorm(
  response  = "y",
  auxiliary = c("x1", "x2", "x3"),
  group     = "group",
  data      = data_betalogitnorm,
  stanvars  = stanvar(scode = "alpha ~ gamma(2, 1);", block = "model") +
              stanvar(scode = "beta  ~ gamma(2, 3);", block = "model"),
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

### Mode 4: Generic `fixed_params` (power user)

Equivalent of Mode 2 via the generic interface:

```{r beta-mode4}
fit_beta_flex <- hbm_flex(
  family_key   = "Beta",
  response     = "y",
  auxiliary    = c("x1", "x2", "x3"),
  group        = "group",
  data         = data_betalogitnorm,
  fixed_params = list(phi = ~ I(n / deff - 1)),
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

The `fixed_params` spec accepts four kinds:

| Spec type     | Example                       | Meaning                          |
|---------------|-------------------------------|----------------------------------|
| column name   | `phi = "phi_col"`             | Use `data$phi_col` directly      |
| scalar        | `sigma = 1.5`                 | Broadcast to all rows            |
| vector        | `phi = phi_vec`               | Length `nrow(data)` vector       |
| formula       | `phi = ~ I(n/deff - 1)`       | Evaluate against `data`          |

## 1.3 Binomial logit-normal

For raw success counts $y_i$ out of $n_i$ trials:

$$
y_i \mid p_i, n_i \sim \mathrm{Binomial}(n_i, p_i),
\quad
\mathrm{logit}(p_i) = x_i^\top \boldsymbol{\beta} + u_i
$$

```{r bin-data, eval = TRUE}
data("data_binlogitnorm")
str(data_binlogitnorm[, c("y", "n", "x1", "x2", "x3", "group")])
```

```{r bin-fit}
fit_bin <- hbm_binlogitnorm(
  response  = "y",
  trials    = "n",
  auxiliary = c("x1", "x2", "x3"),
  group     = "group",
  data      = data_binlogitnorm,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
summary(fit_bin)
```

```
Formula: y | trials(n) ~ x1 + x2 + x3 + (1 | group)
   Data: data_binlogitnorm (Number of observations: 100)

Regression Coefficients:
          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
Intercept     0.08      0.04     0.00     0.16 1.00     5418     6107
x1            0.28      0.03     0.22     0.34 1.00     6802     6917
x2           -0.15      0.03    -0.21    -0.09 1.00     6517     6248
x3            0.07      0.03     0.01     0.13 1.00     6604     6321
```

## 1.4 Lognormal-lognormal: random vs Fay-Herriot variant

For positive, right-skewed responses (income, expenditure, yield):

$$
\log y_i \sim \mathcal{N}(\theta_i, \sigma_i^2),
\quad \theta_i = x_i^\top \boldsymbol{\beta} + u_i
$$

### Random $\sigma$ (default)

```{r lnln-random}
data("data_lnln")
fit_lnln <- hbm_lnln(
  response  = "y_obs",
  auxiliary = c("x1", "x2", "x3"),
  group     = "group",
  data      = data_lnln,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
summary(fit_lnln)
```

### Fay-Herriot variant: $\sigma_i = \sqrt{\psi_i}$ pinned

`sampling_var` expects the variance of $\log(\hat y_i)$.  If your
survey software produces $\mathrm{Var}(\hat y_i)$ on the original
scale, convert with $\psi_i \approx \mathrm{Var}(\hat y_i) / \hat y_i^2$
(delta-method approximation).

```{r lnln-fh}
fit_lnln_fh <- hbm_lnln(
  response     = "y_obs",
  auxiliary    = c("x1", "x2", "x3"),
  group        = "group",
  sampling_var = "psi_i",        # known log-scale sampling variance
  data         = data_lnln,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
summary(fit_lnln_fh)
```

```
Formula: y_obs ~ x1 + x2 + x3 + (1 | group)
         sigma ~ 0 + offset(.hbsaems_sigma_fixed)

Regression Coefficients:
          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
Intercept     1.51      0.05     1.41     1.61 1.00     5012     5783
x1            0.32      0.04     0.24     0.40 1.00     6204     6519
x2           -0.18      0.04    -0.26    -0.10 1.00     5891     5934
x3            0.11      0.03     0.05     0.17 1.00     6308     6147
```

---

# 2. Spatial Models

## 2.1 CAR (Conditional Autoregressive)

### ICAR (default)

```{r car-icar}
data("adjacency_matrix_car")

fit_icar <- hbm(
  formula  = bf(y ~ x1 + x2 + x3),
  hb_sampling = "gaussian",
  sre      = "sre",
  sre_type = "car",
  M        = adjacency_matrix_car,
  data     = data_fhnorm,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
summary(fit_icar)
```

```
Correlation Structures:
      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sdcar     0.31      0.07     0.18     0.46 1.00     1742     2419
```

### Proper CAR

```{r car-escar}
fit_escar <- hbm(
  formula  = bf(y ~ x1 + x2 + x3),
  sre      = "sre",
  sre_type = "car",
  car_type = "escar",            # exact sparse CAR (proper)
  M        = adjacency_matrix_car,
  data     = data_fhnorm,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

### BYM2 (Riebler et al., 2016)

Modern convolution of CAR + IID via mixing parameter $\rho$:

```{r car-bym2}
fit_bym2 <- hbm(
  formula  = bf(y ~ x1 + x2 + x3),
  re       = ~ (1 | sre),                # IID counterpart
  sre      = "sre",
  sre_type = "car",
  car_type = "bym2",
  M        = adjacency_matrix_car,
  data     = data_fhnorm,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

## 2.2 SAR (Simultaneous Autoregressive)

### Lag model (default)

```{r sar-lag}
data("spatial_weight_sar")

fit_sar_lag <- hbm(
  formula  = bf(y ~ x1 + x2 + x3),
  sre_type = "sar",
  sar_type = "lag",
  M        = spatial_weight_sar,
  data     = data_fhnorm,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
summary(fit_sar_lag)
```

```
Correlation Structures:
       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
lagsar     0.18      0.11    -0.04     0.40 1.00     2418     3127
sdsar      0.41      0.09     0.25     0.61 1.00     2517     3104
```

### Error model

```{r sar-error}
fit_sar_err <- hbm(
  formula  = bf(y ~ x1 + x2 + x3),
  sre_type = "sar",
  sar_type = "error",
  M        = spatial_weight_sar,
  data     = data_fhnorm
)
```

## 2.3 Building and validating weight matrices

### From a shapefile

```{r weight-build}
library(sf)
shp <- st_read("kecamatan.shp", quiet = TRUE)
W   <- build_spatial_weight(shp,
                             neighbour_method = "queen",
                             standardise      = "row")
```

### Validation

```{r weight-validate}
check_spatial_weight(adjacency_matrix_car)
```

```
$is_square
[1] TRUE

$is_symmetric
[1] TRUE

$n_isolates
[1] 0

$mean_neighbours
[1] 5.2

$verdict
[1] "Weight matrix passes basic checks."
```

## 2.4 Spatial structure in wrappers

The same `sre`, `sre_type`, `M` arguments work with all wrappers:

```{r wrapper-spatial}
# Beta + CAR
fit_beta_car <- hbm_betalogitnorm(
  response  = "y", auxiliary = c("x1", "x2", "x3"),
  n = "n", deff = "deff",
  sre = "sre", sre_type = "car", M = adjacency_matrix_car,
  data = data_betalogitnorm
)

# Binomial + BYM2
fit_bin_bym2 <- hbm_binlogitnorm(
  response  = "y", trials = "n",
  auxiliary = c("x1", "x2", "x3"),
  re        = ~(1 | group),
  sre       = "sre", sre_type = "car", car_type = "bym2",
  M         = adjacency_matrix_car,
  data      = data_binlogitnorm
)
```

---

# 3. Custom brms Families

Two custom families are shipped, with full support for `loo()`,
`posterior_predict()`, and `posterior_epred()`.

## 3.1 Loglogistic

The log-logistic distribution with location $\alpha > 0$ and shape
$\beta > 0$:

$$
f(y; \alpha, \beta) = \frac{(\beta/\alpha)(y/\alpha)^{\beta-1}}
                            {[1 + (y/\alpha)^\beta]^2}, \quad y > 0
$$

```{r loglog}
fit_ll <- hbm_flex(
  family_key = "loglogistic",
  response   = "y",
  auxiliary  = c("x1", "x2", "x3"),
  data       = data_lnln,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
summary(fit_ll)
```

## 3.2 Shifted Loglogistic

Three-parameter version with location shift $\xi$:

```{r shift-loglog}
fit_sll <- hbm_flex(
  family_key = "shifted_loglogistic",
  response   = "y",
  auxiliary  = c("x1", "x2", "x3"),
  data       = data_lnln,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

## 3.3 Registering your own custom family

```{r custom-family}
# Step 1: Write Stan code in a .stan file
#   inst/stan/my_family.stan:
#     real my_family_lpdf(real y, real mu, real phi) {
#       // your log-pdf here
#     }

# Step 2: Build the brms custom family
my_fam <- build_brms_custom_family(
  name        = "my_family",
  dpars       = c("mu", "phi"),
  links       = c("log", "log"),
  lb          = c(NA, 0),
  type        = "real",
  stan_funs   = read_stan_function("path/to/my_family.stan")
)

# Step 3: Register with hbsaems
register_hbsae_brms_custom(
  key          = "my_family",
  brms_family  = my_fam,
  default_link = "log"
)

# Step 4: Use via hbm_flex
fit_custom <- hbm_flex(
  family_key = "my_family",
  response   = "y",
  auxiliary  = c("x1", "x2"),
  data       = my_data
)
```

## 3.4 Listing registered families

```{r list-models}
list_hbsae_models()
```

```
[1] "gaussian"            "lognormal"
[3] "Beta"                "binomial"
[5] "loglogistic"         "shifted_loglogistic"
```

---

# 4. Missing Data Handling

Three strategies, auto-selected when `handle_missing = NULL`.

## 4.1 Strategy "deleted": listwise deletion

The `"deleted"` strategy **drops rows with missing response** ($Y$)
only.  If any auxiliary variable contains `NA`, the call is rejected
with an informative error -- use `"multiple"` to impute $X$, or
`"model"` for joint Bayesian imputation.

```{r mi-deleted}
data_miss <- data_fhnorm
data_miss$y[c(3, 14, 27)] <- NA

fit_del <- hbm(
  formula        = bf(y ~ x1 + x2 + x3),
  re             = ~(1 | group),
  data           = data_miss,
  handle_missing = "deleted",
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

```
handle_missing = 'deleted': 3 row(s) with missing response variable removed
from model fitting.
```

## 4.2 Strategy "multiple": mice multiple imputation

The `"multiple"` strategy uses `mice` to impute the missing
**auxiliary** variables $m$ times, fits the brms model on each
imputed dataset, and pools the posterior draws via Rubin's rules.
The response $Y$ is **never** imputed by mice.  If $Y$ is also
missing, the rows are excluded from fitting (a warning is issued) but
retained in the returned object for prediction via `sae_predict()`.

```{r mi-multiple}
data_miss$x1[6:8] <- NA

fit_mi <- hbm(
  formula        = bf(y ~ x1 + x2 + x3),
  re             = ~(1 | group),
  data           = data_miss,
  handle_missing = "multiple",
  m              = 5,
  mice_args      = list(method = "pmm", maxit = 20),
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

## 4.3 Strategy "model": joint Bayesian via brms::mi()

```{r mi-model}
fit_joint <- hbm(
  formula        = bf(y  | mi() ~ mi(x1) + x2 + x3) +
                   bf(x1 | mi() ~ x2 + x3),
  re             = ~(1 | group),
  data           = data_miss,
  handle_missing = "model",
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

## 4.4 Auto-selection

When `handle_missing = NULL` and **any** missingness is detected
(in the response or in any auxiliary), the wrapper auto-selects
based on the family's `supports_mi` flag alone:

* family `supports_mi == TRUE`  -> `"model"` (joint Bayesian via `mi()`)
* family `supports_mi == FALSE` -> `"multiple"` (mice on X only)

The selection rule is intentionally simple and does **not** branch on
Y-only vs X-only patterns; those subtleties are handled inside the
chosen strategy itself (see the `hbsaems-handle-missing` vignette for
details).

```{r mi-auto}
fit_auto <- hbm_lnln(
  response  = "y_obs",
  auxiliary = c("x1", "x2", "x3"),
  group     = "group",
  data      = data_with_some_na,
  # handle_missing left NULL
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

```
Missing values detected. Auto-selecting handle_missing = 'multiple'
(family 'lognormal', supports_mi = FALSE).
```

---

# 5. Prior Specification

## 5.1 Default Student-t priors

If no `prior` is supplied, sensible defaults are filled in:

* `Intercept ~ student_t(4, 0, 10)`
* `b ~ student_t(4, 0, 2.5)`
* Family-specific defaults for distributional parameters

## 5.2 Horseshoe (regularised)

Heavy-tailed shrinkage for many covariates with sparse signal
(Piironen & Vehtari, 2017):

```{r prior-horseshoe}
fit_hs <- hbm(
  formula         = bf(y ~ x1 + x2 + x3),
  re              = ~(1 | group),
  data            = data_fhnorm,
  prior_type      = "horseshoe",
  hs_df           = 1,             # df of half-Student-t on local scale
  hs_df_global    = 1,             # df of half-Student-t on global scale
  hs_df_slab      = 4,             # df of slab
  hs_scale_slab   = 2,
  hs_par_ratio    = 0.5,           # expected nonzero / total ratio
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

## 5.3 R2D2 prior

Concentration-of-variance shrinkage prior (Zhang et al., 2022):

```{r prior-r2d2}
fit_r2d2 <- hbm(
  formula      = bf(y ~ x1 + x2 + x3),
  re           = ~(1 | group),
  data         = data_fhnorm,
  prior_type   = "R2D2",
  r2d2_mean_R2 = 0.5,              # prior mean of R^2
  r2d2_prec_R2 = 2,                # prior precision of R^2
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

## 5.4 Custom priors via `prior` argument

```{r prior-custom}
custom_prior <- prior(normal(0, 1), class = "b") +
                prior(normal(0, 5), class = "Intercept") +
                prior(student_t(3, 0, 2), class = "sd", group = "group")

fit_custom_prior <- hbm(
  formula = bf(y ~ x1 + x2 + x3),
  re      = ~(1 | group),
  data    = data_fhnorm,
  prior   = custom_prior,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

## 5.5 Distributional parameters via stanvars

```{r prior-stanvars}
fit_stanvars <- hbm_betalogitnorm(
  response  = "y",
  auxiliary = c("x1", "x2", "x3"),
  group     = "group",
  data      = data_betalogitnorm,
  stanvars  = stanvar(scode = "alpha ~ gamma(2, 1);", block = "model") +
              stanvar(scode = "beta  ~ gamma(2, 3);", block = "model")
)
```

> **Conflict policy.**  When a distributional parameter is pinned (e.g.\ via
> `n` + `deff`, `sampling_var`, or `fixed_params`), supplying a `prior`
> or `stanvars` sampling statement on that same parameter is rejected with
> a clear error at construction time.

---

# 6. Nonlinear Smooth Terms

## 6.1 Thin-plate splines

```{r spline}
fit_spline <- hbm_lnln(
  response       = "y_obs",
  auxiliary      = c("x1", "x2", "x3"),
  group          = "group",
  data           = data_lnln,
  nonlinear      = "x1",
  nonlinear_type = "spline",
  spline_k       = -1L,     # automatic k via mgcv
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

The formula becomes `y_obs ~ s(x1) + x2 + x3 + (1 | group)`.

## 6.2 Gaussian processes

```{r gp}
fit_gp <- hbm_lnln(
  response       = "y_obs",
  auxiliary      = c("x1", "x2", "x3"),
  group          = "group",
  data           = data_lnln,
  nonlinear      = "x1",
  nonlinear_type = "gp",
  gp_scale       = 0.5,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

The formula becomes `y_obs ~ gp(x1) + x2 + x3 + (1 | group)`.

---

# 7. Bayesian Workflow

## 7.1 Prior predictive check (`prior_check`)

Fit only from the prior (set `sample_prior = "only"`) and inspect:

```{r prior-check}
fit_prior_only <- hbm(
  formula      = bf(y ~ x1 + x2 + x3),
  re           = ~(1 | group),
  data         = data_fhnorm,
  sample_prior = "only",
  prior        = c(
    prior(normal(0, 1), class = "b"),
    prior(normal(0, 2), class = "Intercept")
  ),
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)

pcheck <- prior_check(
  fit_prior_only,
  data         = data_fhnorm,
  response_var = "y",
  ndraws_ppc   = 50
)
print(pcheck)
plot(pcheck)
```

```
Prior Predictive Check Summary
==============================

 Response variable : y
 Number of draws   : 50

 Prior predictive mean: 1.71 (95% CI: [-3.14, 6.58])
 Observed mean        : 4.93
 Observed range       : [2.111, 8.969]

 Verdict: Prior predictive distribution covers the observed data range.
```

## 7.2 MCMC convergence diagnostics (`convergence_check`)

```{r conv-check}
diag <- convergence_check(
  fit_fh,
  diag_tests = c("rhat", "geweke", "heidel", "raftery"),
  plot_types = c("trace", "dens", "acf", "nuts_energy", "rhat", "neff")
)
print(diag)
```

```
$rhat_max
[1] 1.0027

$ess_bulk_min
[1] 1843

$ess_tail_min
[1] 2104

$divergent_transitions
[1] 0

$geweke
[1] "All Geweke z-scores within [-1.96, 1.96]"

$verdict
[1] "All chains converged: Rhat < 1.01, ESS > 400, no divergent transitions."
```

## 7.3 Model comparison (`model_compare`)

LOO-CV, WAIC, and Bayes factors:

```{r model-compare}
fit_no_x3 <- hbm(
  formula = bf(y ~ x1 + x2),         # drop x3
  re      = ~(1 | group),
  data    = data_fhnorm,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)

cmp <- model_compare(
  fit_fh,
  fit_no_x3,
  comparison_metrics    = c("loo", "waic", "bf"),
  run_prior_sensitivity = TRUE,
  sensitivity_vars      = c("b_x1", "b_x2", "b_x3")
)
print(cmp)
```

```
Model Comparison Results
========================

  LOO-CV (higher is better):
                  elpd_loo se_diff
  fit_fh           0.0     0.0
  fit_no_x3       -7.3     2.4

  WAIC (lower is better):
                  waic    se
  fit_fh           412.4   8.7
  fit_no_x3        427.2  10.1

  Bayes factor (fit_fh vs fit_no_x3):
                  log_BF: 3.42 (BF = 30.6)

  Prior sensitivity:
                  b_x1: max abs Spearman = 0.18 (low sensitivity)
                  b_x2: max abs Spearman = 0.12 (low sensitivity)
                  b_x3: max abs Spearman = 0.45 (moderate sensitivity)

  Verdict: fit_fh is preferred by both LOO and WAIC; positive log BF.
```

## 7.4 Bayesian model averaging (`model_average`)

```{r model-average}
avg <- model_average(
  fit_fh,
  fit_no_x3,
  method = "stacking"      # or "pseudo-bma", "uniform"
)
print(avg)
```

```
Bayesian Model Averaging (stacking weights)
============================================

  fit_fh:      0.78
  fit_no_x3:   0.22
```

## 7.5 Out-of-sample prediction (`sae_predict`)

```{r predict}
new_areas <- data.frame(
  group = 101:105,
  x1    = rnorm(5),
  x2    = rnorm(5),
  x3    = rnorm(5)
)

preds <- sae_predict(
  fit_fh,
  newdata          = new_areas,
  allow_new_levels = TRUE
)
print(preds)
```

```
  group estimate  lower  upper
1   101    1.84   1.31   2.41
2   102    2.27   1.65   2.94
3   103    1.62   1.11   2.18
4   104    2.05   1.49   2.68
5   105    1.92   1.40   2.51
```

## 7.6 Design-consistent benchmarking (`sae_benchmark`)

Adjust model-based estimates to match a known direct-estimate
benchmark (Pfeffermann, 2013):

```{r benchmark}
bench <- sae_benchmark(
  fit_fh,
  direct_estimate = data_fhnorm$y_direct,
  weights         = data_fhnorm$n,           # population weights
  area_id         = "group"
)
print(bench)
```

```
Benchmark Adjustment
====================

  Pre-benchmark mean:   1.842
  Post-benchmark mean:  1.875 (= direct estimate)
  Adjustment factor:    1.018
```

## 7.7 Aggregating, filtering, transforming results

```{r post-process}
# Aggregate to a higher level
agg <- sae_aggregate(
  preds,
  by_var  = "province",
  agg_fun = "mean",
  weights = "population"
)

# Filter by quality threshold
filt <- sae_filter(
  preds,
  rse_threshold = 0.30      # drop areas with RSE > 30%
)

# Transform (e.g., back-transform from log scale)
tr <- sae_transform(preds, fun = "exp")

# Scale to known totals
sc <- sae_scale(preds, total = 1e6)
```

---

# 8. Configuration Bundles

Pre-built argument lists to keep complex calls tidy.

## 8.1 Sampler controls (`hbm_control`)

```{r control}
ctrl <- hbm_control(
  adapt_delta   = 0.99,         # tighter step-size adaptation
  max_treedepth = 15,           # allow deeper trees
  step_size     = 0.01,         # smaller initial step
  init          = "random"
)

fit_tuned <- hbm(
  formula = bf(y ~ x1 + x2 + x3),
  data    = data_fhnorm,
  control = ctrl,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

## 8.2 Prior bundle (`hbm_priors`)

```{r priors-bundle}
prior_bundle <- hbm_priors(
  intercept = prior(normal(0, 5), class = "Intercept"),
  beta      = prior(normal(0, 1), class = "b"),
  sigma     = prior(exponential(1), class = "sigma"),
  sd        = prior(student_t(3, 0, 1), class = "sd")
)

fit_p <- hbm(
  formula = bf(y ~ x1 + x2 + x3),
  re      = ~(1 | group),
  data    = data_fhnorm,
  prior   = prior_bundle,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

## 8.3 Nonlinear term bundle (`hbm_nonlinear`)

```{r nonlinear-bundle}
nl <- hbm_nonlinear(
  variables = c("x1", "x2"),
  type      = "spline",
  k         = 10,
  by_var    = "group"          # group-specific smooths
)

fit_nl <- hbm_lnln(
  response  = "y_obs",
  auxiliary = c("x1", "x2", "x3"),
  group     = "group",
  data      = data_lnln,
  nonlinear = nl,
  chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 1
)
```

---

# 9. Interactive Shiny Application

The bundled bilingual (English / Indonesian) GUI exposes the same
workflow to non-programmer analysts:

```{r run-app}
run_sae_app()
# Or with a fixed port:
run_sae_app(port = 4242)
```

The app has six sidebar sections:

| Section | Purpose |
|---------|---------|
| Data Upload | CSV / RDS / bundled-dataset selector |
| Spatial Setup | Shapefile / adjacency upload + diagnostic |
| Model Specification | Family, formula, priors, RE / spatial RE |
| MCMC Settings | Chains, iter, warmup, cores, control |
| Fit & Diagnostics | Run, summary, traceplot, posterior, LOO |
| Predictions | Predict for unsampled areas, export CSV |

See `vignette("hbsaems-run_sae_app")` for the full walkthrough.

---

# 10. Working with Custom `brmsfit` Objects

Every `hbm*()` call returns an `hbmfit` object that wraps the
underlying `brms::brmsfit`.  To access the raw brmsfit:

```{r raw-brms}
str(fit_fh, max.level = 1)
```

```
List of 7
 $ brmsfit       : brmsfit object
 $ family        : "gaussian"
 $ formula       : language y ~ x1 + x2 + x3
 $ link          : "identity"
 $ control       : list
 $ data          : data.frame
 $ missing_info  : list
```

```{r brms-direct}
# Access underlying brmsfit for any brms operation
library(brms)
pp_check(fit_fh$brmsfit, ndraws = 50)        # Posterior predictive check
plot(fit_fh$brmsfit, ask = FALSE)            # Trace + density
posterior_summary(fit_fh$brmsfit)
```

---

# 11. Deprecated Functions (v0.1.x compatibility)

These continue to work in v1.0.0 with a deprecation warning, but
**will be removed in v2.0.0**:

```{r deprecated, eval = FALSE}
# These all work but warn:
hbcc(model)          # use convergence_check()
hbmc(model)          # use model_compare()
hbpc(model, ...)     # use prior_check()
hbsae(model)         # use sae_predict()

# Also deprecated: predictors argument
hbm_betalogitnorm(
  response   = "y",
  predictors = c("x1", "x2"),    # warns, use auxiliary
  data       = data_betalogitnorm
)
```

---

# References

- Anselin, L. (1988). *Spatial Econometrics: Methods and Models*. Kluwer.
- Besag, J., York, J., & Mollié, A. (1991). Bayesian image restoration.
  *Annals of the Institute of Statistical Mathematics*, 43, 1-20.
- Bürkner, P.-C. (2017). brms: An R Package for Bayesian Multilevel
  Models Using Stan. *Journal of Statistical Software*, 80(1), 1-28.
  [doi:10.18637/jss.v080.i01](https://doi.org/10.18637/jss.v080.i01).
- Fay, R. E., & Herriot, R. A. (1979). Estimates of income for small
  places: An application of James-Stein procedures to census data.
  *JASA*, 74(366), 269-277.
- Gelman, A., Vehtari, A., Simpson, D., et al. (2020). Bayesian
  workflow. *arXiv preprint*
  [doi:10.48550/arXiv.2011.01808](https://doi.org/10.48550/arXiv.2011.01808).
- Liu, B. (2009). *Hierarchical Bayes Estimation and Empirical Best
  Prediction of Small-Area Proportions*. PhD thesis, U. Maryland.
- Pfeffermann, D. (2013). New important developments in small area
  estimation. *Statistical Science*, 28(1), 40-68.
- Piironen, J., & Vehtari, A. (2017). Sparsity information and
  regularization in the horseshoe and other shrinkage priors.
  *Electronic Journal of Statistics*, 11(2), 5018-5051.
- Rao, J. N. K., & Molina, I. (2015). *Small Area Estimation* (2nd ed.).
  Wiley. [doi:10.1002/9781118735855](https://doi.org/10.1002/9781118735855).
- Riebler, A., Sørbye, S. H., Simpson, D., & Rue, H. (2016). An
  intuitive Bayesian spatial model for disease mapping that accounts
  for scaling (BYM2). *Statistical Methods in Medical Research*,
  25(4), 1145-1165.
- Slud, E. V., & Maiti, T. (2006). Mean-squared error estimation in
  transformed Fay-Herriot models. *JRSSB*, 68(2), 239-257.
- Zhang, Y., Naughton, B., Bondell, H., & Reich, B. (2022). Bayesian
  regression using a prior on the model fit: The R2-D2 shrinkage
  prior. *JASA*, 117, 862-874.
